---
title: "Midterm Challenge: Philadelphia Housing Price Prediction"
subtitle: "MUSA 5080 - Public Policy Analytics"
format: 
  html:
    toc: true
    toc-location: left
    code-fold: show
    theme: cosmo
---

## Overview

**Due Date:** Mar 16, 2026, 11:59PM

**In-Class Presentations:** March 17, 2026 (5 minutes per team)

**Weight:** 15% of final grade

**Team:** You'll work with your table-mates as a team. Feel free to delegate. Everyone should upload their final products onto their own portfolio websites. Be sure to acknowledge your team-mates.

**Submission Format:**

1. **Presentation Slides** (.qmd → revealjs, ~10-15 slides) - Main deliverable (see my weekly lecture notes for inspiration on how to make slides!)
2. **Technical Appendix** (.qmd → HTML document) - Supporting details

---

## The Challenge

You are consultancy (please name your consultancy) competing to win the bid to work for a project for the **Philadelphia Office of Property Assessment**. The city wants to improve its Automated Valuation Model (AVM) for property tax assessments. Your task is to build a predictive model for residential sale prices and **present your findings** to city officials in a 5-minute briefing. 

**Deliverables:**

1. **Presentation slides** (10-15 slides MAX) - Your main findings for stakeholders
2. **Technical appendix** (HTML document) - All code, diagnostics, and detailed analysis
3. **5-minute in-class presentation** - Deliver your slides on March 17

**Your goal:** Predict 2023-2024 home sale prices accurately while communicating findings clearly to a policy audience.

---

## Two-Part Submission

### Part A: Presentation Slides (Primary Deliverable)

**Format:** Quarto revealjs presentation

**Example YAML:**
```yaml
---
title: "Philadelphia Housing Price Prediction"
subtitle: "Improving Property Tax Assessments"
author: "Your Name"
format: 
  revealjs:
    theme: simple
    slide-number: true
    smaller: true
---
```

**Content:** ~10-15 (could be less!!) slides covering:
1. Research question & motivation (1-2 slides)
2. Data overview (1 slide)
3. Key visualizations (2-3 slides)
4. Model comparison results (1-2 slides)
5. Main findings (2-3 slides)
6. Policy recommendations (1-2 slides)

**Audience:** City officials who don't know R or care about the nitty gritty of stats. They just want the best estimates possible. 

**No code** in these slides - just polished visualizations and key takeaways

---

### Part B: Technical Appendix (Supporting Documentation)

**Format:** Quarto HTML document

**Example YAML:**
```yaml
---
title: "Philadelphia Housing Model - Technical Appendix"
author: "Your Name"
format: 
  html:
    code-fold: show
    toc: true
    toc-location: left
    theme: cosmo
---
```

**Content:** All the technical details:
- Complete data cleaning code
- All EDA visualizations
- Feature engineering code
- Full model outputs
- Diagnostic plots
- Detailed interpretations

**Audience:** Data scientists and technical reviewers

**All code visible** - this is where you show your work

---

## Data Sources

### Primary Dataset: Philadelphia Property Sales

**Source:** [Philadelphia Property Sales](https://metadata.phila.gov/#home/datasetdetails/5543865f20583086178c4ee5/representationdetails/55d624fdad35c7e854cb21a4/)

This dataset contains actual property sales with:

- Sale price
- Sale date
- Property characteristics (bedrooms, bathrooms, sq ft, etc.)
- Property location (address, coordinates)

**You will need to:**

- Download the data
- Clean it (missing values, outliers, data errors)
- Filter to 2023-2025 residential sales only



```{r packages}
echo = FALSE
library(sf)
library(tidyverse)
library(tidycensus)
library(tigris)
library(MASS)
library(dplyr)
library(scales)
library(ggplot2)
library(caret)
library(car)
library(knitr)
library(readr)
library(patchwork)
```
```{r load api}
echo = FALSE
census_api_key("fadfb87c0c7417766780c9bf52b8df9adef61e21", overwrite = TRUE)

Sys.getenv("CENSUS_API_KEY")
```

```{r primary data}
phl_props <- st_read("C:/Users/Liz/Desktop/y2s2/PPA/crouse/labs/lab_3/data/opa_properties_public.geojson")

phl_props_filtered <- phl_props %>% 
  filter( category_code_description %in% c("SINGLE FAMILY", "MULTI FAMILY"),
    year(as.Date(sale_date)) %in% c(2023, 2024),
    mailing_city_state == "PHILADELPHIA PA",
    !is.na(sale_price) & sale_price > 50000,
    !is.na(total_livable_area) & total_livable_area > 0,
    !is.na(number_of_bedrooms) & number_of_bedrooms > 0
  )
phl_props_filtered <- phl_props_filtered %>% st_transform(2722)
```


### Secondary Datasets (You Choose!)

**Required:** Browse the OpenPhily Data portal and use Census Data to incorporate spatial features into your model.

**Your task:** Think like an urban planner. What location factors matter for housing prices in Philadelphia?

```{r secondary data}
variables <- c(med_inc = "B19013_001", 
              burden_renter_total = "B25070_001",
              burden_renter_30 = "B25070_007",
              burden_renter_35 = "B25070_008",
              burden_renter_40 = "B25070_009",
              burden_renter_50 = "B25070_010",
              burden_owner_total = "B25091_001",
              burden_owner_30 = "B25091_008",
              burden_owner_35 = "B25091_009",
              burden_owner_40 = "B25091_010",
              burden_owner_50 = "B25091_011",
              poverty_total = "B17001_001",
              poverty_below = "B17001_002",
              total_households = "B11001_001",
              single_family = "B25024_002",
              single_family_att = "B25024_003",
              multi_family = "B25024_004",
              multi_family_3_4 = "B25024_005",
              multi_family_5_9 = "B25024_006",
              multi_family_10_19 = "B25024_007",
              multi_family_20_49 = "B25024_008",
              multi_family_50plus = "B25024_009",
              edu_total = "B15003_001",
              edu_bachelor = "B15003_022",
              edu_master = "B15003_023",
              edu_professional = "B15003_024",
              edu_phd = "B15003_025",
              commute_total = "B08301_001",
              commute_car = "B08301_002",
              commute_transit = "B08301_010")

acs_data <- get_acs(
  geography = "tract",
  variables = variables,
  state = "PA",
  county = "Philadelphia",
  year = 2023,
  survey = "acs5",
  geometry = TRUE,
  output = "wide")

acs_data <- acs_data %>% st_transform(2722)

props_tract <- st_join(phl_props_filtered, acs_data)
prop_summary <- props_tract %>%
  st_drop_geometry() %>%
  group_by(GEOID) %>%
  summarise(
    mean_price = mean(sale_price, na.rm = TRUE),
    n_properties = n()
  )

tract_analysis <- acs_data %>%
  left_join(prop_summary, by = "GEOID")
  
```
---

## Assignment Structure

Your work should follow this workflow, with results split between presentation and appendix:

### Phase 1: Data Preparation (Technical Appendix)

**Load and clean Philadelphia sales data:**

- Filter to residential properties, 2023-2024 sales
- Remove obvious errors
- Handle missing values
- Document all cleaning decisions

**Load secondary data:**

- Census data (tidycensus):
- Spatial amenities (OpenDataPhilly)
- Join to sales data appropriately
- Make sure you have the correct CRS!

**Deliverable (Appendix only):**

- Complete data cleaning code
- Summary tables showing before/after dimensions
- Narrative explaining decisions

---

### Phase 2: Exploratory Data Analysis

**Create at least 5 professional visualizations:**

1. Distribution of sale prices (histogram)

```{r plot1}
phl_props_filtered <- phl_props_filtered %>%
  mutate(log_sale_price = log(sale_price))

ggplot(phl_props_filtered, aes(x = log_sale_price)) +
  geom_histogram(
    bins = 30,
    color = "black",
    fill = "steelblue"
  ) +
  labs(
    title = "Distribution of Log Sale Prices",
    x = "Log Sale Price",
    y = "Number of Properties"
  ) +
  theme_minimal()

```  

2. Geographic distribution (map)
```{r plot2}
ggplot(phl_props_filtered) +
  geom_sf(aes(color = log_sale_price), size = .001) +
  scale_color_viridis_c(option = "plasma") +
  theme_void() +
  labs(title = "Geographic Distribution of Sale Prices", color = "Sale Price ($)")
```
3. Price vs. structural features (scatter plots)
```{r plot3}
ggplot(phl_props_filtered,
       aes(x = number_of_bedrooms, y = sale_price)) +
  geom_point(alpha = 0.6, size = 1.5) +
  scale_y_continuous(labels = scales::comma) +
  theme_minimal() +
  labs(
    title = "Sale Price vs. Number of Bedrooms",
    x = "Number of Bedrooms",
    y = "Sale Price ($)"
  )
```
4. Price vs. spatial features (scatter plots)
**see feature engineering section 


5. One creative visualization

```{r plot5}
ggplot(tract_analysis) +
  geom_sf(aes(fill = n_properties)) +
  scale_fill_viridis_c() +
  theme_void() +
  labs(
    title = "Properties Sold by Tract",
    fill = "Number of Properties"
  )

```


**For presentation slides:** Select your **best 2-3 visualizations** that tell a compelling story

**For appendix:** Include all visualizations with detailed interpretations

**Example presentation slide:**

```markdown
## Where Are Expensive Homes in Philadelphia?

[Beautiful map showing price patterns]

**Key Findings:**
- Center City and University City command premium prices
- River wards show emerging appreciation
- Northeast Philadelphia remains most affordable
```

---

### Phase 3: Feature Engineering (Technical Appendix)

**Create spatial features: (these are examples below, but how you construct your model is up to your team)**

1. **Buffer-based features**:

   - Parks within 500ft, 1000ft
   - Transit stops within 400ft
   - Schools, crime, etc.
   
```{r buffers}

##For our buffer spatial feature, we chose to count the number of trees within a 500 foot radius of each property, since tree cover often correlates to property value. 

phl_trees <- st_read("C:/Users/Liz/Desktop/y2s2/PPA/crouse/labs/lab_3/data/ppr_tree_inventory_2025.geojson")
phl_trees <- phl_trees %>% st_transform(2722)

##Create property buffers
prop_buffers_500 <- st_buffer(phl_props_filtered, dist = 500)

trees_props <- st_join(
  phl_trees,
  prop_buffers_500,
  join = st_within,
  left = FALSE)
trees_by_prop <- trees_props %>%
  group_by(pin) %>%
  summarise(
    n_trees = n(),
    .groups = "drop"
  )
##Join back to buffers
prop_buffers_trees <- prop_buffers_500 %>%
  left_join(
    st_drop_geometry(trees_by_prop),
    by = "pin") %>%
  mutate(n_trees = replace_na(n_trees, 0)
  )


```



##** TEAM NOTE: I can't render the following code for this plot #4 in phase 2 bc I only just create tree features here. If someone can think of a way to do it or change it so it can go earlier that would be great


```{r spatial plot}

ggplot(prop_buffers_trees,
       aes(x = n_trees, y = sale_price)) +
  geom_point(alpha = 0.6, size = 1.5) +
  scale_y_continuous(labels = scales::comma) +
  theme_minimal() +
  labs(
    title = "Sale Price vs. Number of Trees in 500ft Radius",
    x = "Number of Trees",
    y = "Sale Price ($)"
  )
   
```

2. **k-Nearest Neighbor features**:

   - Average distance to k nearest parks, transit, etc.
   
   
```{r k nearest neighbor}   

##For k-nearest neighbor, we chose to evaluate the distance of each property to the nearest 3, 5, and 10 trees, as this often correlates with property values. 

schools <- st_read("C:/Users/Liz/Desktop/y2s2/PPA/crouse/labs/lab_3/data/Schools")
schools <- schools %>% st_transform(2722)

##Creating the distance matrix

school_dist_matrix <- st_distance(phl_props_filtered, schools)
school_dist_matrix <- as.matrix(school_dist_matrix)
school_dist_matrix <- units::drop_units(school_dist_matrix)



#Performing knn function
school_knn <- function(school_dist_matrix, k) {
  apply(school_dist_matrix, 1, function(x) {
    mean(sort(x)[1:k])
  })
}

# run function
school_knn_3 <- school_knn(school_dist_matrix, k = 3)
school_knn_5 <- school_knn(school_dist_matrix, k = 5)
school_knn_10 <- school_knn(school_dist_matrix, k = 10)

# attach to dataset
phl_props_filtered$school_knn_3 <- school_knn_3
phl_props_filtered$school_knn_5 <- school_knn_5
phl_props_filtered$school_knn_10 <- school_knn_10

##Evaluating results

phl_props_filtered %>%
  st_drop_geometry() %>%
  dplyr::select(school_knn_3) %>%
  summary()

phl_props_filtered %>%
  st_drop_geometry() %>%
  dplyr::select(school_knn_5) %>%
  summary()

phl_props_filtered %>%
  st_drop_geometry() %>%
  dplyr::select(school_knn_10) %>%
  summary()

```

3. **Census variables**:


4. **Interaction terms**:



**Deliverable (Appendix only):**

- All feature engineering code
- Summary table of features created
- Brief justification for each feature

---

### Phase 4: Model Building

**Build models progressively: (for example)**

1. Structural features only
2. + Census variables
3. + Spatial features
4. + Interactions and fixed effects

**For presentation slides:** Show **one comparison table** (RMSE, R² for 4 different models you constructed in your process)

**For appendix:**

- Complete model code
- Full modelsummary output
- Coefficient interpretations

**Example presentation slide:**

```markdown
## Model Performance Improves with Each Layer

| Model | CV RMSE (log) | R² |
|-------|---------------|-----|
| Structural Only | 0.42 | 0.61 |
| + Census | 0.38 | 0.69 |
| + Spatial | 0.31 | 0.78 |
| + Interactions/FE | 0.26 | 0.84 |

**Bottom line:** Neighborhood effects matter most!
```

---

### Phase 5: Model Validation

**Use 10-fold cross-validation:**
- Compare all 4 models
- Report RMSE, MAE, R² for each
- Create predicted vs. actual plot

**For presentation slides:** Final CV results table (shown above) + one compelling visual

**For appendix:** 

- Complete CV code
- Detailed results
- Predicted vs. actual scatter plot
- Discussion of which features matter most

---

### Phase 6: Model Diagnostics (Technical Appendix Only)

**Check assumptions for best model:**

- Residual plot (linearity, homoscedasticity)
- Q-Q plot (normality)
- Cook's distance (influential observations)

**Deliverable (Appendix only):**

- All 3 diagnostic plots
- Interpretation of each
- How you addressed violations (if any)

**Note:** Don't include diagnostic plots in presentation - too technical!

---

### Phase 7: Conclusions & Recommendations

**Answer these questions:**

1. What is your final model's accuracy?
2. Which features matter most for Philadelphia prices?
3. Which neighborhoods are hardest to predict?
4. Equity concerns?
5. Limitations?

**For presentation slides:** 1-2 slides with clear, concise answers (bullet points)

**For appendix:** 2-3 paragraphs with detailed discussion

**Example presentation slide:**

```markdown
## Key Findings & Recommendations

**Model Accuracy:** RMSE = 0.26 (log scale) ≈ 26% typical error

**Top Predictors:**
- Neighborhood fixed effects (largest impact)
- Square footage (β = 0.0003, p < 0.001)
- Distance to transit (β = -0.05, p < 0.001)

**Recommendations:**
✓ Current AVM undervalues transit-accessible properties  
✓ Model struggles in rapidly gentrifying neighborhoods  
```

---

## Submission Requirements

### What to Submit (by 11:59 PM, March 16, 2026)

**Upload to Canvas - A link to your portfolio that Contains**

1. **Presentation Slides**

   - `LastName_FirstName_Presentation.html` (rendered slides)
   - `LastName_FirstName_Presentation.qmd` (source file)
   - Must use `format: revealjs`
   - 10-15 slides maximum
   - No code visible in slides

2. **Technical Appendix**

   - `LastName_FirstName_Appendix.html` (rendered document)
   - `LastName_FirstName_Appendix.qmd` (source file)
   - Must use `format: html`
   - All code visible and commented
   - Complete analysis documented

3. **Data files** OR clear download instructions in appendix

**For Teams:** Use `LastName1_LastName2_Presentation.html`

### In-Class Presentation (March 17, 2026)

**Format:** 5 minutes per team

**What to present:**

- Walk through your presentation slides. Choose your team's spoke's person or take turns You'll all stand up there and try to look calm, confident, & collected.
- Hit the highlights - research question, key viz, model results, recommendations
- Speak to a policy audience (your classmates are pretending to be city officials)
- Be ready for 1-2 questions
- You are trying to win the bid! Convince the audience of your agency's work.

**What NOT to do:**

- Don't read slides verbatim
- Don't show code
- Don't go into technical details
- Don't go over 5 minutes (I'll cut you off!)

---

## Grading Rubric (Scaled to 15% of course grade)

### Presentation Slides 

| Component | Points | Criteria |
|-----------|--------|----------|
| **Research Question** | 2 | Clear motivation, Set the stage |
| **Data Overview** | 2 | Concise description of sources, sample size |
| **Visualizations** | 3 | 2-3 polished, publication-quality visualizations; clear takeaways |
| **Model Comparison** | 3 | Clean results table; clear winner; interprets improvement |
| **Key Findings** | 3 | Top predictors identified; coefficients interpreted correctly |
| **Presentation Quality** | 3 | Professional design, no typos, flows logically, appropriate for audience |

**Key:** Slides should tell a compelling story without technical jargon. Imagine presenting to the Deputy Mayor.

---

### In-Class Presentation

| Component | Points | Criteria |
|-----------|--------|----------|
| **Content** | 2 | Covers key points efficiently, answers questions thoughtfully |
| **Time Management** | 2 | Finishes within 5 minutes without rushing |

---

### Technical Appendix 

| Component | Points | Criteria |
|-----------|--------|----------|
| **Data Cleaning** | 3 | Complete code, proper filtering, missing value handling, documentation |
| **EDA** | 3 | 5+ visualizations, each with interpretation |
| **Feature Engineering** | 3 | Buffers , kNN , census , interactions , all properly created |
| **Model Building** | 3 | 4 progressive models, proper specification, handles sparse categories |
| **Cross-Validation** | 3 | Proper 10-fold CV, results table, code runs without errors |
| **Diagnostics** | 3 | Residual plots included, interpreted, violations addressed |
| **Code Quality** | 3 | Clean, commented, reproducible, follows best practices, no errors |

**Key:** This is where technical reviewers verify your work. All code must run without errors & be reproducible!!.

---


## Example Presentation Structure

**Slide 1: Title**
- Your team name and teammates
- Project title
- Date

**Slide 2: The Problem**


**Slide 3: Data Sources**
- Property sales (n = X,XXX, 2023-2024)
- Census ACS (income, education, poverty)
- OpenDataPhilly (parks, transit, crime)

**Slide 4: Where Are Expensive Homes?**
- [Map visualization]
- Key pattern observed

**Slide 5: What Drives Prices?**
- [Best scatter plot or faceted visualization]
- Key relationship identified

**Slide 6: Model Comparison**
- [Results table]
- "Each layer improves prediction"

**Slide 7: Top Predictors**
- Neighborhood (biggest impact)
- Square footage (β = X)
- Transit access (β = Y)

**Slide 8: Model Performance**
- Final RMSE: 0.26 (log scale)
- Translation: ~26% typical error
- Beats baseline by 40%

**Slide 9: Hardest to Predict**
- [Visualization of residuals by neighborhood]


**Slide 10: Recommendations**


**Slide 11: Limitations & Next Steps**


**Slide 12: Questions?**
- Thank you
- [Contact info]

---

## Tips for Success

### Start Early
- Data cleaning always takes longer than expected
- OpenDataPhilly can be slow to download
- Leave time for troubleshooting AND RENDERING!!!

### Check Your Work

- Organize your file directory from the beginning. 
- Run your entire .qmd file from scratch with a clean workspace before submitting 
- Make sure all visualizations display
- Check that your narrative flows logically



---

## Frequently Asked Questions

**Q: Do I have to create both slides AND an appendix?**  

A: Yes! Slides are your main deliverable (present findings). Appendix proves you did the work correctly.

**Q: Can code appear in my presentation slides?** 

A: NO! Slides are for city officials. All code goes in the technical appendix.

**Q: How many slides should I have?**  

A: 10-15 maximum. Quality over quantity. Each slide should have a clear purpose.

**Q: Can I use a different city?**  
A: No, everyone uses Philadelphia for comparability.

**Q: How do I make my .qmd render as revealjs slides?**  
A: Use `format: revealjs` in your YAML (see template above). Test it early! See
the .qmd files behind my lecture slides for examples.

**Q: My presentation is 8 minutes. Is that okay?**  
A: NO! You must cut it to 5 minutes. Practice and trim ruthlessly.

**Q: Should I include all 5 EDA visualizations in my slides?**  
A: No! Slides should have your best 2-3 visualizations. Put all 5 in the appendix.

**Q: My RMSE is 0.35 in log scale. Is that good?**  
A: Depends on your data, but 0.25-0.45 is typical for hedonic models. Compare to your baseline or put your data back into dollars!

**Q: Should I remove all outliers?**  
A: No! Only remove obvious errors. Use log transformations to handle legitimate outliers.

**Q: What if my code works on my computer but not when I render it?**  
A: Start fresh, restart R, render in a clean session. Check for hard-coded paths. Be
aware that sometimes quarto will point to an error on one line, but actually the
error is somewhere else.

**Q: Can I use ChatGPT/Claude to write my analysis?**  
A: You may use AI for debugging code, but NOT for writing your analysis. I want to
read your interpretation in your own words.

**Q: How formal should my presentation be?**  
A: Professional but not stuffy. Like you're briefing a city council member who's smart but doesn't know statistics.

**Q: What happens if I go over 5 minutes?**  
A: I'll politely cut you off and you'll lose points. Practice with a timer!

---

## Example of Your Group's Workflow

**Week 1:**
- Download data
- Initial cleaning
- Basic EDA

**Week 2:**
- Feature engineering
- Build 4 models
- Run cross-validation
- Diagnostics
- Write conclusions
- Proofread and submit

---

## Academic Integrity

- You may discuss concepts with classmates
- You may NOT share code or slides with other teams
- All work must be your own (or your teams)
- Cite any external resources used
- Acknowledge how you used AI in your work and which AI you used (For example, Claude helped me today in coming up with a draft of this assignment, but I edited it thoroughly!)

---

## Final Checklist Before Submitting

### Presentation Slides
- [ ] Renders correctly as revealjs slides
- [ ] 10-15 slides maximum
- [ ] No code visible anywhere
- [ ] 2-3 polished visualizations
- [ ] Clear model comparison table
- [ ] No typos or formatting errors
- [ ] Professional theme applied
- [ ] Tested presentation timing (<5 minutes)

### Technical Appendix
- [ ] Renders correctly as HTML document
- [ ] All code visible and commented
- [ ] Data cleaning fully documented
- [ ] 5+ EDA visualizations included
- [ ] All features properly created
- [ ] 4 models with full output
- [ ] 10-fold CV code runs without errors
- [ ] Diagnostic plots included
- [ ] Sparse categories handled correctly
- [ ] Tested that it knits without errors

### Both Files
- [ ] Proper file naming convention
- [ ] Both .qmd source files included
- [ ] Data sources clearly documented
- [ ] No hard-coded file paths (use relative paths! - this should be reproducible)
- [ ] Uploaded to Canvas on time

---
