---
title: "Lab 4: Spatial Predictive Analysis"
subtitle: "Analyzing 311 Violations"
author: "Liz Crouse"
date: April 6, 2026
format: 
  html:
    code-fold: false
    toc: true
    toc-location: left
    theme: cosmo
    embed-resources: true
execute:
  warning: false
  message: false
---

```{r setup}

#| message: false
#| warning: false
#| echo: false


# Load required packages
library(tidyverse)      # Data manipulation
library(sf)             # Spatial operations
library(here)           # Relative file paths
library(viridis)        # Color scales
library(terra)          # Raster operations (replaces 'raster')
library(spdep)          # Spatial dependence
library(FNN)            # Fast nearest neighbors
library(MASS)           # Negative binomial regression
library(patchwork)      # Plot composition (replaces grid/gridExtra)
library(knitr)          # Tables
library(kableExtra)     # Table formatting
library(classInt)       # Classification intervals


# Spatstat for analyzing point patterns - split into sub-packages
library(spatstat.geom)    # Spatial geometries
library(spatstat.explore) # Spatial exploration/KDE

# Set options
options(scipen = 999)  # No scientific notation
set.seed(5080)         # Reproducibility, could be any number

# Create consistent theme for visualizations
theme_crime <- function(base_size = 11) {
  theme_minimal(base_size = base_size) +
    theme(
      plot.title = element_text(face = "bold", size = base_size + 1),
      plot.subtitle = element_text(color = "gray30", size = base_size - 1),
      legend.position = "right",
      panel.grid.minor = element_blank(),
      axis.text = element_blank(),
      axis.title = element_blank()
    )
}

# Set as default
theme_set(theme_crime())

cat("✓ All packages loaded successfully!\n")
cat("✓ Working directory:", getwd(), "\n")

```

#### Part 1: Data Loading & Exploration

-   Load your 311 data and Chicago spatial boundaries
-   Create visualizations showing the spatial distribution of your violation type
-   Describe patterns you observe

```{r boundaries}

#| message: false

# Load police districts (used for spatial cross-validation)
policeDistricts <- 
  st_read("https://data.cityofchicago.org/api/geospatial/24zt-jpfn?method=export&format=GeoJSON") %>%
  st_transform('ESRI:102271') %>%
  dplyr::select(District = dist_num)

# Load police beats (smaller administrative units)
policeBeats <- 
  st_read("https://data.cityofchicago.org/api/geospatial/n9it-hstw?method=export&format=GeoJSON") %>%
  st_transform('ESRI:102271') %>%
  dplyr::select(Beat = beat_num)

# Load Chicago boundary
chicagoBoundary <- 
  st_read("https://raw.githubusercontent.com/urbanSpatial/Public-Policy-Analytics-Landing/master/DATA/Chapter5/chicagoBoundary.geojson") %>%
  st_transform('ESRI:102271')

cat("✓ Loaded spatial boundaries\n")
cat("  - Police districts:", nrow(policeDistricts), "\n")
cat("  - Police beats:", nrow(policeBeats), "\n")


```

```{r pothole data}
#| echo: FALSE
potholes <- read.csv("C:/Users/lizmc/Desktop/Y2S2/ppa/crouse/labs/lab_4/data/311_Service_Requests_20260328.csv")

potholes <- potholes[!is.na(potholes$LONGITUDE) & !is.na(potholes$LATITUDE), ]
potholes <- st_as_sf(potholes, coords = c("LONGITUDE", "LATITUDE"), crs = 4326)
potholes <- potholes %>%
  st_transform('ESRI:102271')

cat("  - Number of potholes:", nrow(potholes), "/n")
cat("  - CRS:", st_crs(potholes)$input, "\n")
cat("  - Date range:", format(min(potholes$CREATED_DATE, na.rm = TRUE)), "to", 
    format(max(potholes$CREATED_DATE, na.rm = TRUE)), "\n")

```

```{r visualizations}
#| fig-width: 10
#| fig-height: 5

# Simple point map
p1 <- ggplot() + 
  geom_sf(data = chicagoBoundary, fill = "gray95", color = "gray60") +
  geom_sf(data = potholes, color = "#d62828", size = 0.1, alpha = 0.4) +
  labs(
    title = "Pothole Locations",
    subtitle = paste0("Chicago 2024-2025, n = ", nrow(potholes))
  )

# Density surface using modern syntax
p2 <- ggplot() + 
  geom_sf(data = chicagoBoundary, fill = "gray95", color = "gray60") +
  geom_density_2d_filled(
    data = data.frame(st_coordinates(potholes)),
    aes(X, Y),
    alpha = 0.7,
    bins = 8
  ) +
  scale_fill_viridis_d(
    option = "plasma",
    direction = -1,
    guide = "none"  
  ) +
  labs(
    title = "Density Surface",
    subtitle = "Kernel density estimation"
  )

# Combine plots using patchwork 
p1 + p2 + 
  plot_annotation(
    title = "Spatial Distribution of Potholes in Chicago",
    tag_levels = 'A'
  )

# Describe patterns you observe
## Potholes are concentrated along the eastern edge of Chicago (shoreline of Lake Michigan), specifically in the northeastern area of the city. Smaller clusters appear in the north and southwest corners.
```

#### Part 2: Fishnet Grid Creation

-   Create a 500m x 500m fishnet grid
-   Aggregate your violations to grid cells
-   Visualize the count distribution

```{r fishnet}
# Create 500m x 500m grid
fishnet <- st_make_grid(
  chicagoBoundary,
  cellsize = 500,  # 500 meters per cell
  square = TRUE
) %>%
  st_sf() %>%
  mutate(uniqueID = row_number()) %>%
  st_transform('ESRI:102271') 

# Keep only cells that intersect Chicago
fishnet <- fishnet[chicagoBoundary, ]

# View basic info
cat("✓ Created fishnet grid\n")
cat("  - Number of cells:", nrow(fishnet), "\n")
cat("  - Cell size:", 500, "x", 500, "meters\n")
cat("  - Cell area:", round(st_area(fishnet[1,])), "square meters\n")
```

```{r aggregate}
# Spatial join: which cell contains each burglary?
potholes_fishnet <- st_join(potholes, fishnet, join = st_within) %>%
  st_drop_geometry() %>%
  group_by(uniqueID) %>%
  summarize(countPotholes = n()) 


glimpse(potholes_fishnet)


# Join back to fishnet (cells with 0 burglaries will be NA)
fishnet <- fishnet %>%
  left_join(potholes_fishnet, by = "uniqueID") %>%
  mutate(countPotholes = replace_na(countPotholes, 0))
```

```{r visualize aggregate}
#| fig-width: 8
#| fig-height: 6

# Visualize aggregated counts
ggplot() +
  geom_sf(data = fishnet, aes(fill = countPotholes), color = NA) +
  geom_sf(data = chicagoBoundary, fill = NA, color = "NA", linewidth = 1) +
  scale_fill_viridis_c(
    name = "Potholes",
    option = "plasma",
    trans = "sqrt",  # Square root for better visualization of skewed data
    breaks = c(0, 100, 200, 300, 400)
  ) +
  labs(
    title = "Potholes Counts by Grid Cell",
    subtitle = "500m x 500m cells, Chicago 2024-2025"
  ) +
  theme_crime()
```

#### Part 3: Spatial Features

-   Calculate k-nearest neighbor features
-   Perform Local Moran's I analysis
-   Identify hot spots and cold spots
-   Create distance-to-hotspot measures
-   Join any additional contextual data if you are looking for more to do and really get into this (e.g., demographics, land use)

```{r kernel density}
#| message: false

# Convert potholes to ppp (point pattern) format for spatstat
potholes_ppp <- as.ppp(
  st_coordinates(potholes),
  W = as.owin(st_bbox(chicagoBoundary))
)

# Calculate KDE with 1km bandwidth
kde_potholes <- density.ppp(
  potholes_ppp,
  sigma = 1000,  # 1km bandwidth
  edge = TRUE    # Edge correction
)

# Convert to terra raster (modern approach, not raster::raster)
kde_raster <- rast(kde_potholes)

# Extract KDE values to fishnet cells
fishnet <- fishnet %>%
  mutate(
    kde_value = terra::extract(
      kde_raster,
      vect(fishnet),
      fun = mean,
      na.rm = TRUE
    )[, 2]  # Extract just the values column
  )

cat("✓ Calculated KDE baseline\n")
```

```{r visualize kernel density}
#| fig-width: 8
#| fig-height: 6

ggplot() +
  geom_sf(data = fishnet, aes(fill = kde_value), color = NA) +
  geom_sf(data = chicagoBoundary, fill = NA, color = "white", linewidth = 1) +
  scale_fill_viridis_c(
    name = "KDE Value",
    option = "plasma"
  ) +
  labs(
    title = "Kernel Density Estimation Baseline",
    subtitle = "Simple spatial smoothing of pothole locations"
  ) +
  theme_crime()
```

```{r knn}
#| message: false

# Get coordinates
fishnet_coords <- st_coordinates(st_centroid(fishnet))
pothole_coords <- st_coordinates(potholes)

# Calculate k nearest neighbors and distances
nn_result <- get.knnx(pothole_coords, fishnet_coords, k = 3)

# Add to fishnet
fishnet <- fishnet %>%
  mutate(
    potholes.nn = rowMeans(nn_result$nn.dist)
  )

cat("✓ Calculated nearest neighbor distances\n")
summary(fishnet$potholes.nn)
```

```{r morans}

# Function to calculate Local Moran's I
calculate_local_morans <- function(data, variable, k = 5) {
  
  # Create spatial weights
  coords <- st_coordinates(st_centroid(data))
  neighbors <- knn2nb(knearneigh(coords, k = k))
  weights <- nb2listw(neighbors, style = "W", zero.policy = TRUE)
  
  # Calculate Local Moran's I
  local_moran <- localmoran(data[[variable]], weights)
  
  # Classify clusters
  mean_val <- mean(data[[variable]], na.rm = TRUE)
  
  data %>%
    mutate(
      local_i = local_moran[, 1],
      p_value = local_moran[, 5],
      is_significant = p_value < 0.05,
      
      moran_class = case_when(
        !is_significant ~ "Not Significant",
        local_i > 0 & .data[[variable]] > mean_val ~ "High-High",
        local_i > 0 & .data[[variable]] <= mean_val ~ "Low-Low",
        local_i < 0 & .data[[variable]] > mean_val ~ "High-Low",
        local_i < 0 & .data[[variable]] <= mean_val ~ "Low-High",
        TRUE ~ "Not Significant"
      )
    )
}

# Apply to potholes
fishnet <- calculate_local_morans(fishnet, "countPotholes", k = 5)
```

```{r visualize morans}
#| fig-width: 8
#| fig-height: 6

# Visualize hot spots
ggplot() +
  geom_sf(
    data = fishnet, 
    aes(fill = moran_class), 
    color = NA
  ) +
  scale_fill_manual(
    values = c(
      "High-High" = "#d7191c",
      "High-Low" = "#fdae61",
      "Low-High" = "#abd9e9",
      "Low-Low" = "#2c7bb6",
      "Not Significant" = "gray90"
    ),
    name = "Cluster Type"
  ) +
  labs(
    title = "Local Moran's I: Potholes Clusters",
    subtitle = "High-High = Hot spots of disorder"
  ) +
  theme_crime()
```

```{r distance to hotspots}
# Get centroids of "High-High" cells (hot spots)
hotspots <- fishnet %>%
  filter(moran_class == "High-High") %>%
  st_centroid()

# Calculate distance from each cell to nearest hot spot
if (nrow(hotspots) > 0) {
  fishnet <- fishnet %>%
    mutate(
      dist_to_hotspot = as.numeric(
        st_distance(st_centroid(fishnet), hotspots %>% st_union())
      )
    )
  
  cat("✓ Calculated distance to abandoned car hot spots\n")
  cat("  - Number of hot spot cells:", nrow(hotspots), "\n")
} else {
  fishnet <- fishnet %>%
    mutate(dist_to_hotspot = 0)
  cat("⚠ No significant hot spots found\n")
}
```

#### Part 4: Count Regression Models

-   Fit Poisson regression
-   Fit Negative Binomial regression
-   Compare model fit (AIC)

```{r join police districts}
# Strip District if it exists from a previous run
fishnet <- fishnet %>% dplyr::select(-any_of("District"))

# Join district information to fishnet via centroids
fishnet_centroids <- st_join(
  st_centroid(fishnet),
  dplyr::select(policeDistricts, District),
  join = st_within,
  left = TRUE
) %>%
  st_drop_geometry() %>%
  dplyr::select(uniqueID, District)

fishnet <- fishnet %>%
  left_join(fishnet_centroids, by = "uniqueID")

cat("✓ Joined police districts\n")
cat("  - Districts:", length(unique(fishnet$District)), "\n")
cat("  - Cells with District:", sum(!is.na(fishnet$District)), "\n")
```

```{r clean dataset for poisson}
# Create clean modeling dataset
fishnet_model <- fishnet %>%
  st_drop_geometry() %>%
  dplyr::select(
    uniqueID,
    District,
   countPotholes,
    potholes.nn,
    dist_to_hotspot
  ) %>%
  na.omit()  # Remove any remaining NAs

cat("✓ Prepared modeling data\n")
cat("  - Observations:", nrow(fishnet_model), "\n")
cat("  - Variables:", ncol(fishnet_model), "\n")
```

```{r fit poisson}
# Fit Poisson regression
model_poisson <- glm(
  countPotholes ~ potholes.nn + dist_to_hotspot,
  data = fishnet_model,
  family = "poisson"
)

# Summary
summary(model_poisson)
```

```{r binomial regression}
# Fit Negative Binomial model
model_nb <- glm.nb(
  countPotholes ~ potholes.nn + 
    dist_to_hotspot,
  data = fishnet_model
)

# Summary
summary(model_nb)

# Compare AIC (lower is better)
cat("\nModel Comparison:\n")
cat("Poisson AIC:", round(AIC(model_poisson), 1), "\n")
cat("Negative Binomial AIC:", round(AIC(model_nb), 1), "\n")

```

#### Part 5: Spatial Cross-Validation

-   Implement Leave-One-Group-Out cross-validation on data
-   Calculate and report error metrics (MAE, RMSE)

```{r spatial cross validation}

# Get unique districts
districts <- unique(fishnet_model$District)
cv_results <- tibble()

cat("Running LOGO Cross-Validation...\n")

for (i in seq_along(districts)) {
  
  test_district <- districts[i]
  
  # Split data
  train_data <- fishnet_model %>% filter(District != test_district)
  test_data <- fishnet_model %>% filter(District == test_district)
  
  # Fit model on training data
  model_cv <- glm.nb(
  countPotholes ~ potholes.nn + dist_to_hotspot,
    data = train_data
  )
  
  # Predict on test data
  test_data <- test_data %>%
    mutate(
      prediction = predict(model_cv, test_data, type = "response")
    )
  
  # Calculate metrics
  mae <- mean(abs(test_data$countPotholes - test_data$prediction))
  rmse <- sqrt(mean((test_data$countPotholes - test_data$prediction)^2))
  
  # Store results
  cv_results <- bind_rows(
    cv_results,
    tibble(
      fold = i,
      test_district = test_district,
      n_test = nrow(test_data),
      mae = mae,
      rmse = rmse
    )
  )
  
  cat("  Fold", i, "/", length(districts), "- District", test_district, 
      "- MAE:", round(mae, 2), "\n")
}

# Overall results
cat("\n✓ Cross-Validation Complete\n")
cat("Mean MAE:", round(mean(cv_results$mae), 2), "\n")
cat("Mean RMSE:", round(mean(cv_results$rmse), 2), "\n")
```

```{r final predictions}
# Fit final model on all data
final_model <- glm.nb(
   countPotholes ~ potholes.nn + dist_to_hotspot,
  data = fishnet_model
)

# Add predictions back to fishnet
fishnet <- fishnet %>%
  mutate(
    prediction_nb = predict(final_model, fishnet_model, type = "response")[match(uniqueID, fishnet_model$uniqueID)]
  )

# Also add KDE predictions (normalize to same scale as counts)
kde_sum <- sum(fishnet$kde_value, na.rm = TRUE)
count_sum <- sum(fishnet$countPotholes, na.rm = TRUE)
fishnet <- fishnet %>%
  mutate(
    prediction_kde = (kde_value / kde_sum) * count_sum
  )
```

#### Part 6: Model Evaluation

-   Compare to KDE baseline

```{r compare to baseline}
#| fig-width: 12
#| fig-height: 4

# Create three maps
p1 <- ggplot() +
  geom_sf(data = fishnet, aes(fill = countPotholes), color = NA) +
  scale_fill_viridis_c(name = "Count", option = "plasma", limits = c(0, 15)) +
  labs(title = "Actual Potholes") +
  theme_crime()

p2 <- ggplot() +
  geom_sf(data = fishnet, aes(fill = prediction_nb), color = NA) +
  scale_fill_viridis_c(name = "Predicted", option = "plasma", limits = c(0, 15)) +
  labs(title = "Model Predictions (Neg. Binomial)") +
  theme_crime()

p3 <- ggplot() +
  geom_sf(data = fishnet, aes(fill = prediction_kde), color = NA) +
  scale_fill_viridis_c(name = "Predicted", option = "plasma", limits = c(0, 15)) +
  labs(title = "KDE Baseline Predictions") +
  theme_crime()

p1 + p2 + p3 +
  plot_annotation(
    title = "Actual vs. Predicted Potholes",
    subtitle = "Does our complex model outperform simple KDE?"
  )
```

```{r comparison metrics}
# Calculate performance metrics
comparison <- fishnet %>%
  st_drop_geometry() %>%
  filter(!is.na(prediction_nb), !is.na(prediction_kde)) %>%
  summarize(
    model_mae = mean(abs(countPotholes - prediction_nb)),
    model_rmse = sqrt(mean((countPotholes - prediction_nb)^2)),
    kde_mae = mean(abs(countPotholes - prediction_kde)),
    kde_rmse = sqrt(mean((countPotholes - prediction_kde)^2))
  )

comparison %>%
  pivot_longer(everything(), names_to = "metric", values_to = "value") %>%
  separate(metric, into = c("approach", "metric"), sep = "_") %>%
  pivot_wider(names_from = metric, values_from = value) %>%
  kable(
    digits = 2,
    caption = "Model Performance Comparison"
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```
