---
title: "Lab 2: Spatial Analysis and Visualization"
subtitle: "Healthcare Access and Equity in Pennsylvania"
author: "Liz Crouse"
date: 2/13/26
format: 
  html:
    code-fold: false
    toc: true
    toc-location: left
    theme: cosmo
    embed-resources: true
execute:
  warning: false
  message: false
---

## Assignment Overview

**Learning Objectives:**
- Apply spatial operations to answer policy-relevant research questions
- Integrate census demographic data with spatial analysis
- Create publication-quality visualizations and maps
- Work with spatial data from multiple sources
- Communicate findings effectively for policy audiences

---

## Part 1: Healthcare Access for Vulnerable Populations

### Research Question

**Which Pennsylvania counties have the highest proportion of vulnerable populations (elderly + low-income) living far from hospitals?**

Your analysis should identify counties that should be priorities for healthcare investment and policy intervention.

### Required Analysis Steps

Complete the following analysis, documenting each step with code and brief explanations:

#### Step 1: Data Collection (5 points)

Load the required spatial data:
- Pennsylvania county boundaries
- Pennsylvania hospitals (from lecture data)
- Pennsylvania census tracts

**Your Task:**
```{r}
echo = FALSE

# Load required packages
library(tidycensus)
library(tidyverse)
library(stringr)
library(ggplot2)
library(dplyr)
library(tigris)
library(sf)
library(viridis)
library(RColorBrewer)
library(knitr)
library(scales)


# Load spatial data
pa_hospitals <- st_read("C:/Users/lizmc/Desktop/Y2S2/ppa/crouse/labs/lab_2/data/hospitals.geojson")
pa_counties <- st_read("C:/Users/lizmc/Desktop/Y2S2/ppa/crouse/labs/lab_2/data/Pennsylvania_County_Boundaries")
pa_tracts <- tracts("PA")


# Check that all data loaded correctly
plot(pa_hospitals)
plot(pa_counties)
plot(pa_tracts)
st_crs(pa_hospitals)
st_crs(pa_counties)
st_crs(pa_tracts)

```

**Questions to answer:**
- How many hospitals are in your dataset? **223**
- How many census tracts? **3,446**
- What coordinate reference system is each dataset in? **Hospitals and counties are in WGS 84; tracts are in NAD83.**

---

#### Step 2: Get Demographic Data 

Use `tidycensus` to download tract-level demographic data for Pennsylvania.

**Required variables:**
- Total population
- Median household income
- Population 65 years and over (you may need to sum multiple age categories)

**Your Task:**
```{r}
# Get demographic data from ACS
variables <- c(mhi = "B19013_001",
               pop = "B01003_001",
               m65_66 = "B01001_020",
    m67_69 = "B01001_021",
    m70_74 = "B01001_022",
    m75_79 = "B01001_023",
    m80_84 = "B01001_024",
    m85p   = "B01001_025",

    f65_66 = "B01001_044",
    f67_69 = "B01001_045",
    f70_74 = "B01001_046",
    f75_79 = "B01001_047",
    f80_84 = "B01001_048",
    f85p   = "B01001_049"
  )


tract_demographics <- get_acs(
  geography = "tract",
  state = "PA",
  variables = variables,
  year = 2022,
  survey = "acs5",
  output = "wide"
  
)
tract_demographics <- tract_demographics %>%
  mutate(
    pop_65plus =
      (m65_66E + m67_69E + m70_74E + m75_79E + m80_84E + m85pE +
      f65_66E + f67_69E + f70_74E + f75_79E + f80_84E + f85pE),

    pct_65plus = (pop_65plus / popE) * 100
  )


# Join to tract boundaries
tract_demographics <- left_join(tract_demographics, pa_tracts, by = "GEOID")

#Initial calculations
sum(is.na(tract_demographics$mhiE))
mean(tract_demographics$mhiE, na.rm = TRUE)

```

**Questions to answer:**
- What year of ACS data are you using? **5-Year ACS 2022**
- How many tracts have missing income data? **63**
- What is the median income across all PA census tracts? **$77,527.23***

---

#### Step 3: Define Vulnerable Populations 

Identify census tracts with vulnerable populations based on TWO criteria:
1. Low median household income (choose an appropriate threshold) 
2. Significant elderly population (choose an appropriate threshold) 

**Your Task:**
```{r}
# Filter for vulnerable tracts based on your criteria
vulnerable_tracts <- tract_demographics %>%
  filter(mhiE < 64300 & pct_65plus > 25)
old_tracts <- tract_demographics %>%
  filter(pct_65plus > 25)
lowinc_tracts <- tract_demographics %>%
  filter(mhiE < 64300)

```

**Questions to answer:**
- What income threshold did you choose and why? I chose $64,300 as my low-income threshold because it is 200% of the PA poverty line for a household of four. 200% poverty line is often used as a metric for eligibility for social services, which would likely coincide with the tracts that need additional health services. This represents about 40% of tracts.
- What elderly population threshold did you choose and why? I chose 25% because that means at least one in four individuals in the tract are 65+, meaning extra health services may be needed.
- How many tracts meet your vulnerability criteria? **230**
- What percentage of PA census tracts are considered vulnerable by your definition? 6.7%

---

#### Step 4: Calculate Distance to Hospitals 

For each vulnerable tract, calculate the distance to the nearest hospital.

**Your Task:**
```{r}
# Transform to appropriate projected CRS
st_geometry(vulnerable_tracts) <- "geometry"
vulnerable_tracts <- vulnerable_tracts %>%
  st_transform(crs = 2272)
pa_hospitals <- pa_hospitals %>%
  st_transform(crs = 2272)
# Calculate distance from each tract centroid to nearest hospital
tract_centroids <- st_centroid(vulnerable_tracts)
vulnerable_tracts <- vulnerable_tracts %>%
  mutate(
    nearest_hosp = st_distance(
      tract_centroids, pa_hospitals$geometry[st_nearest_feature(tract_centroids, pa_hospitals)],
      by_element = TRUE
    )
    
  )
vulnerable_tracts <- vulnerable_tracts %>%
  mutate(
    nearest_hosp_miles = as.numeric(nearest_hosp) / 1609.34
  )
mean(vulnerable_tracts$nearest_hosp_miles, na.rm = TRUE)
tracts_15p <- vulnerable_tracts %>%
  filter(nearest_hosp_miles > 15)
```

**Requirements:**
- Use an appropriate projected coordinate system for Pennsylvania
- Calculate distances in miles
- Explain why you chose your projection **I used EPSG 2272 because it is a project coordinate system specific to the southern part of Pennsylvania, meaning it will provide more accurate measurements.

**Questions to answer:**
- What is the average distance to the nearest hospital for vulnerable tracts? **18.64 mi**
- What is the maximum distance? **95.66 mi**
- How many vulnerable tracts are more than 15 miles from the nearest hospital? **90**

---

#### Step 5: Identify Underserved Areas 

Define "underserved" as vulnerable tracts that are more than 15 miles from the nearest hospital.

**Your Task:**
```{r}
# Create underserved variable
vulnerable_tracts <- vulnerable_tracts %>%
  mutate(service_level = case_when(nearest_hosp_miles > 15 ~ "underserved",
            TRUE ~ "served"))
vulnerable_tracts %>%
  count(service_level)



```

**Questions to answer:**
- How many tracts are underserved? **90**
- What percentage of vulnerable tracts are underserved? **39%**
- Does this surprise you? Why or why not? **This does not surprise me. Lower income tracts with high elderly populations are often more rural and therefore may have less access to services. If anything, I am surprised it is not a larger proportion.**

---

#### Step 6: Aggregate to County Level

Use spatial joins and aggregation to calculate county-level statistics about vulnerable populations and hospital access.

**Your Task:**
```{r}
# Spatial join tracts to counties
pa_counties <- st_transform(pa_counties, st_crs(vulnerable_tracts))
county_statistics <- st_join(vulnerable_tracts, pa_counties["COUNTY_NAM"])

# Aggregate statistics by county

county_statistics_summary <- county_statistics %>%
  group_by(COUNTY_NAM) %>%
  summarise(
    n_vuln_tracts = n(),
    n_unders_tracts = sum(service_level == "underserved", na.rm = TRUE),
    underserved_pct = (n_unders_tracts / n_vuln_tracts) * 100,
    avg_dist_nearest_hosp = mean(nearest_hosp_miles, na.rm = TRUE),
    total_pop_vuln = sum(popE, na.rm = TRUE))

underserved_counties <- county_statistics_summary %>%
  filter(underserved_pct == 100)


countygeo_stat_summary <- pa_counties %>%
  st_join(county_stat_summary)


```

**Required county-level statistics:**
- Number of vulnerable tracts
- Number of underserved tracts  
- Percentage of vulnerable tracts that are underserved
- Average distance to nearest hospital for vulnerable tracts
- Total vulnerable population

**Questions to answer:**
- Which 5 counties have the highest percentage of underserved vulnerable tracts? **By my parameters for under-service, there are 17 counties in PA that are entirely underserved. The five with the least service (furthest average distance to nearest hospital) are Dauphin, Perry, Sullivan, Bradford, and Pike Counties.**
- Which counties have the most vulnerable people living far from hospitals? **Counties that are 100% underserved with populations greater than 10,000 are Wayne, McKean, Pike, Tioga, and Clarion Counties.**
- Are there any patterns in where underserved counties are located? **The largest concentration of the most severely underserved tracts is in north central PA, a rural area distant from PA's two major population centers of Pittsburgh and Philadelphia.**

---

#### Step 7: Create Summary Table 

Create a professional table showing the top 10 priority counties for healthcare investment.

**Your Task:**
```{r}
# Create and format priority counties table


underserved_counties_nogeom <- underserved_counties %>%
  st_drop_geometry() %>%
  arrange(desc(avg_dist_nearest_hosp)) %>%
  slice(1:10)
  
underserved_table <- underserved_counties_nogeom %>%
   mutate(
    n_vuln_tracts = comma(n_vuln_tracts),
    n_unders_tracts = comma(n_unders_tracts),
    total_pop_vuln = comma(total_pop_vuln),
    underserved_pct = percent(underserved_pct / 100, accuracy = 0.1),
    avg_dist_nearest_hosp = round(avg_dist_nearest_hosp, 1)) %>%
  kable(col.names = c("County","# Vulnerable Tracts","# Underserved Tracts","% Tracts Underserved","Average Distance from Tract to Nearest Hospital", "Total Vulnerable Population"), caption = "Counties with high populations of elderly and low income individuals with inequitable access to hospitals")
view(underserved_table)

```

**Requirements:**
- Use `knitr::kable()` or similar for formatting
- Include descriptive column names
- Format numbers appropriately (commas for population, percentages, etc.)
- Add an informative caption
- Sort by priority (you decide the metric)

---

## Part 2: Comprehensive Visualization 

Using the skills from Week 3 (Data Visualization), create publication-quality maps and charts.

### Map 1: County-Level Choropleth 

Create a choropleth map showing healthcare access challenges at the county level.

**Your Task:**
```{r}
# Create county-level access map

ggplot(countygeo_stat_summary) + 
  geom_sf(aes(fill = underserved_pct), color = "black") +
  scale_fill_viridis_b(
    option = "mako",
    direction = -1,
    name = 
      "% Underserved") +
  theme_void() + 
  labs(title = "Percentage of Underserved Tracts by County")





```

**Requirements:**
- Fill counties by percentage of vulnerable tracts that are underserved
- Include hospital locations as points
- Use an appropriate color scheme
- Include clear title, subtitle, and caption
- Use `theme_void()` or similar clean theme
- Add a legend with formatted labels

---

### Map 2: Detailed Vulnerability Map 

Create a map highlighting underserved vulnerable tracts.

**Your Task:**
```{r}
# Create detailed tract-level map
pa_tracts <- pa_tracts %>%
  st_transform(2272)

tracts_service_level <- pa_tracts %>%
  st_join(vulnerable_tracts["service_level"], join = st_within)
tracts_service_level <- tracts_service_level %>%
  mutate(service_level = ifelse(is.na(service_level), "served", service_level))
tracts_service_level %>% 
  count(service_level)


ggplot() + 
  geom_sf(
    data = tracts_service_level,
    aes(fill = service_level),
    color = NA) +
  geom_sf(
    data = pa_counties,
    fill = NA,
    color = "black",
    size = 0.2
  ) +
  scale_fill_manual(
    values = c(
      "underserved" = "red",
      "served" = NA),
    na.value = NA) +
  theme_void() +
  labs(
    title = "Vulnerable & Underserved Tracts in PA Counties", subtitle = "Tracts with low median incomes, high 65+ population, and >15 miles to nearest hospital",
    fill = "Service Level")



```

**Requirements:**
- Show underserved vulnerable tracts in a contrasting color
- Include county boundaries for context
- Show hospital locations
- Use appropriate visual hierarchy (what should stand out?)
- Include informative title and subtitle

---

### Chart: Distribution Analysis

Create a visualization showing the distribution of distances to hospitals for vulnerable populations.

**Your Task:**
```{r}
# Create distribution visualization
ggplot(data = vulnerable_tracts, aes(x = nearest_hosp_miles)) + 
  geom_histogram(binwidth = 5, color = "black") +
    scale_x_continuous(
    limits = c(0, NA),
    expand = c(0, 0)
  ) +
  scale_y_continuous(
    limits = c(0, NA),
    expand = c(0, 0)) +
     labs(title = "Distribution of distances to nearest hospital among vulnerable PA census tracts",
     x = "Distance to nearest hospital (mi)",
     y = "Number of tracts",
     caption = "The nearest hospital for over 90 vulnerable tracts is more than 15 miles away.")



```

**Suggested chart types:**
- Histogram or density plot of distances
- Box plot comparing distances across regions
- Bar chart of underserved tracts by county
- Scatter plot of distance vs. vulnerable population size

**Requirements:**
- Clear axes labels with units
- Appropriate title
- Professional formatting
- Brief interpretation (1-2 sentences as a caption or in text)

---

## Part 3: Bring Your Own Data Analysis 

Choose your own additional spatial dataset and conduct a supplementary analysis.

### Challenge Options

Choose ONE of the following challenge exercises, or propose your own research question using OpenDataPhilly data (https://opendataphilly.org/datasets/). 

**Note these are just loose suggestions to spark ideas - follow or make your own as the data permits and as your ideas evolve. This analysis should include bringing in your own dataset, ensuring the projection/CRS of your layers align and are appropriate for the analysis (not lat/long or geodetic coordinate systems). The analysis portion should include some combination of spatial and attribute operations to answer a relatively straightforward question**

---

#### Education & Youth Services

**Option A: Educational Desert Analysis**
- **Data:** Schools, Libraries, Recreation Centers, Census tracts (child population)
- **Question:** "Which neighborhoods lack adequate educational infrastructure for children?"
- **Operations:** Buffer schools/libraries (0.5 mile walking distance), identify coverage gaps, overlay with child population density
- **Policy relevance:** School district planning, library placement, after-school program siting

**Option B: School Safety Zones**
- **Data:** Schools, Crime Incidents, Bike Network
- **Question:** "Are school zones safe for walking/biking, or are they crime hotspots?"
- **Operations:** Buffer schools (1000ft safety zone), spatial join with crime incidents, assess bike infrastructure coverage
- **Policy relevance:** Safe Routes to School programs, crossing guard placement

---

#### Environmental Justice

**Option C: Green Space Equity** 
- **Data:** Parks, Street Trees, Census tracts (race/income demographics)
- **Question:** "Do low-income and minority neighborhoods have equitable access to green space?"
- **Operations:** Buffer parks (10-minute walk = 0.5 mile), calculate tree canopy or park acreage per capita, compare by demographics
- **Policy relevance:** Climate resilience, environmental justice, urban forestry investment
---

#### Public Safety & Justice

**Option D: Crime & Community Resources**
- **Data:** Crime Incidents, Recreation Centers, Libraries, Street Lights
- **Question:** "Are high-crime areas underserved by community resources?"
- **Operations:** Aggregate crime counts to census tracts or neighborhoods, count community resources per area, spatial correlation analysis
- **Policy relevance:** Community investment, violence prevention strategies
---

#### Infrastructure & Services

**Option E: Polling Place Accessibility**
- **Data:** Polling Places, SEPTA stops, Census tracts (elderly population, disability rates)
- **Question:** "Are polling places accessible for elderly and disabled voters?"
- **Operations:** Buffer polling places and transit stops, identify vulnerable populations, find areas lacking access
- **Policy relevance:** Voting rights, election infrastructure, ADA compliance

---

#### Health & Wellness

**Option F: Recreation & Population Health**
- **Data:** Recreation Centers, Playgrounds, Parks, Census tracts (demographics)
- **Question:** "Is lack of recreation access associated with vulnerable populations?"
- **Operations:** Calculate recreation facilities per capita by neighborhood, buffer facilities for walking access, overlay with demographic indicators
- **Policy relevance:** Public health investment, recreation programming, obesity prevention

---

#### Emergency Services

**Option G: EMS Response Coverage**
- **Data:** Fire Stations, EMS stations, Population density, High-rise buildings
- **Question:** "Are population-dense areas adequately covered by emergency services?"
- **Operations:** Create service area buffers (5-minute drive = ~2 miles), assess population coverage, identify gaps in high-density areas
- **Policy relevance:** Emergency preparedness, station siting decisions

---

#### Arts & Culture

**Option H: Cultural Asset Distribution**
- **Data:** Public Art, Museums, Historic sites/markers, Neighborhoods
- **Question:** "Do all neighborhoods have equitable access to cultural amenities?"
- **Operations:** Count cultural assets per neighborhood, normalize by population, compare distribution across demographic groups
- **Policy relevance:** Cultural equity, tourism, quality of life, neighborhood identity

---

### Data Sources

**OpenDataPhilly:** https://opendataphilly.org/datasets/
- Most datasets available as GeoJSON, Shapefile, or CSV with coordinates
- Always check the Metadata for a data dictionary of the fields.

**Additional Sources:**
- **Pennsylvania Open Data:** https://data.pa.gov/
- **Census Bureau (via tidycensus):** Demographics, economic indicators, commute patterns
- **TIGER/Line (via tigris):** Geographic boundaries

### Recommended Starting Points

**If you're feeling confident:** Choose an advanced challenge with multiple data layers. 
**If you are a beginner, choose something more manageable that helps you understand the basics**


**If you have a different idea:** Propose your own question! Just make sure:
- You can access the spatial data
- You can perform at least 2 spatial operations

### Your Analysis

**Your Task:**

1. **Find and load additional data**
   - Document your data source
   - Check and standardize the CRS
   - Provide basic summary statistics

```{r}
# Load your additional dataset




```

**Questions to answer:**
- What dataset did you choose and why?
- What is the data source and date?
- How many features does it contain?
- What CRS is it in? Did you need to transform it?

---

2. **Pose a research question**

Write a clear research statement that your analysis will answer.

**Examples:**
- "Do vulnerable tracts have adequate public transit access to hospitals?"
- "Are EMS stations appropriately located near vulnerable populations?"
- "Do areas with low vehicle access have worse hospital access?"

---

3. **Conduct spatial analysis**

Use at least TWO spatial operations to answer your research question.

**Required operations (choose 2+):**
- Buffers
- Spatial joins
- Spatial filtering with predicates
- Distance calculations
- Intersections or unions
- Point-in-polygon aggregation

**Your Task:**
```{r}
# Your spatial analysis










```

**Analysis requirements:**
- Clear code comments explaining each step
- Appropriate CRS transformations
- Summary statistics or counts
- At least one map showing your findings
- Brief interpretation of results (3-5 sentences)

**Your interpretation:**

[Write your findings here]


## Finally - A few comments about your incorporation of feedback!
Take a few moments to clean up your markdown document and then write a line or two or three about how you may have incorporated feedback that you recieved after your first assignment. 

---

## Submission Requirements

**What to submit:**

1. **Rendered HTML document posted to your course portfolio** with all code, outputs, maps, and text
   - Use `embed-resources: true` in YAML so it's a single file
   - All code should run without errors
   - All maps and charts should display correctly
2. Submit the correct and working links of your assignment on Canvas

---

