
#Part 1
Adapt this code to work with your quarter:

Update date ranges for weather data

Check for any data structure changes

Create the same 5 models

Calculate MAE for each model

```{r setup}
#| message: false
#| warning: false

library(tidyverse)
library(lubridate)
library(sf)
library(tigris)
library(tidycensus)
library(riem)

library(viridis)
library(knitr)
library(kableExtra)
library(patchwork)


library(here)
options(scipen = 999)
```

```{r theme}
plotTheme <- theme_minimal() +
  theme(
    plot.title    = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 10),
    plot.caption  = element_text(size = 8),
    axis.text.x   = element_text(size = 10, angle = 45, hjust = 1),
    axis.text.y   = element_text(size = 10),
    axis.title    = element_text(size = 11, face = "bold"),
    panel.grid.major = element_line(colour = "#D0D0D0", linewidth = 0.2),
    panel.grid.minor = element_blank(),
    axis.ticks    = element_blank(),
    legend.position = "right"
  )

mapTheme <- theme_void() +
  theme(
    plot.title    = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 10),
    plot.caption  = element_text(size = 8),
    legend.position  = "right",
    plot.margin      = margin(1, 1, 1, 1, "cm"),
    legend.key.height = unit(1, "cm"),
    legend.key.width  = unit(0.2, "cm")
  )

palette5 <- c("#eff3ff", "#bdd7e7", "#6baed6", "#3182bd", "#08519c")
```

```{r set census key}
census_api_key(Sys.getenv("CENSUS_API_KEY"))
```
```{r data setup}
indego_q3 <- read_csv(here("labs/lab_5/data/indego-trips-2025-q3.csv"))
indego_q1 <- read_csv(here("labs/lab_5/data/indego-trips-2025-q1.csv"))

```
```{r data exploration}
# How many trips?
cat("Total trips in Q3 2025:", nrow(indego_q3), "\n")


# Date range
cat("Date range:", 
    min(mdy_hm(indego_q3$start_time)), "to", 
    max(mdy_hm(indego_q3$start_time)), "\n")


# How many unique stations?
cat("Unique start stations:", length(unique(indego_q3$start_station)), "\n")


# Trip types
table(indego_q3$trip_route_category)

# Passholder types
table(indego_q3$passholder_type)

# Bike types
table(indego_q3$bike_type)
```
```{r create bins}
indego_q3 <- indego_q3 %>%
  mutate(
    # Parse datetime
    start_datetime = mdy_hm(start_time),
    end_datetime = mdy_hm(end_time),
    
    # Create hourly bins
    interval60 = floor_date(start_datetime, unit = "hour"),
    
    # Extract time features
    week = week(interval60),
    month = month(interval60, label = TRUE),
    dotw = wday(interval60, label = TRUE),
    hour = hour(interval60),
    date = as.Date(interval60),
    
    # Create useful indicators
    weekend = ifelse(dotw %in% c("Sat", "Sun"), 1, 0),
    rush_hour = ifelse(hour %in% c(7, 8, 9, 16, 17, 18), 1, 0)
  )

# Look at temporal features
head(indego_q3 %>% select(start_datetime, interval60, week, dotw, hour, weekend))

indego_q1 <- indego_q1 %>%
  mutate(
    # Parse datetime
    start_datetime = mdy_hm(start_time),
    end_datetime = mdy_hm(end_time),
    
    # Create hourly bins
    interval60 = floor_date(start_datetime, unit = "hour"),
    
    # Extract time features
    week = week(interval60),
    month = month(interval60, label = TRUE),
    dotw = wday(interval60, label = TRUE),
    hour = hour(interval60),
    date = as.Date(interval60),
    
    # Create useful indicators
    weekend = ifelse(dotw %in% c("Sat", "Sun"), 1, 0),
    rush_hour = ifelse(hour %in% c(7, 8, 9, 16, 17, 18), 1, 0)
  )

# Look at temporal features
head(indego_q1 %>% select(start_datetime, interval60, week, dotw, hour, weekend))


```


```{r trips over time}
#| message: false
#| warning: false

# Daily trip counts
daily_trips_q3 <- indego_q3 %>%
  group_by(date) %>%
  summarize(trips = n())

ggplot(daily_trips_q3, aes(x = date, y = trips)) +
  geom_line(color = "#3182bd", linewidth = 1) +
  geom_smooth(se = FALSE, color = "red", linetype = "dashed") +
  labs(
    title = "Indego Daily Ridership - Q3 2025",
    subtitle = "Winter demand patterns in Philadelphia",
    x = "Date",
    y = "Daily Trips",
    caption = "Source: Indego bike share"
  ) +
  plotTheme

# Daily trip counts
daily_trips_q1 <- indego_q1 %>%
  group_by(date) %>%
  summarize(trips = n())

ggplot(daily_trips_q1, aes(x = date, y = trips)) +
  geom_line(color = "#3182bd", linewidth = 1) +
  geom_smooth(se = FALSE, color = "red", linetype = "dashed") +
  labs(
    title = "Indego Daily Ridership - Q3 2025",
    subtitle = "Winter demand patterns in Philadelphia",
    x = "Date",
    y = "Daily Trips",
    caption = "Source: Indego bike share"
  ) +
  plotTheme
```

```{r time patterns}
# Average trips by hour and day type
hourly_patterns_q3 <- indego_q3 %>%
  group_by(hour, weekend) %>%
  summarize(avg_trips = n() / n_distinct(date)) %>%
  mutate(day_type = ifelse(weekend == 1, "Weekend", "Weekday"))

ggplot(hourly_patterns_q3, aes(x = hour, y = avg_trips, color = day_type)) +
  geom_line(linewidth = 1.2) +
  scale_color_manual(values = c("Weekday" = "#08519c", "Weekend" = "#6baed6")) +
  labs(
    title = "Average Hourly Ridership Patterns Q3",
    subtitle = "Clear commute patterns on weekdays",
    x = "Hour of Day",
    y = "Average Trips per Hour",
    color = "Day Type"
  ) +
  plotTheme

hourly_patterns_q1 <- indego_q1 %>%
  group_by(hour, weekend) %>%
  summarize(avg_trips = n() / n_distinct(date)) %>%
  mutate(day_type = ifelse(weekend == 1, "Weekend", "Weekday"))

ggplot(hourly_patterns_q1, aes(x = hour, y = avg_trips, color = day_type)) +
  geom_line(linewidth = 1.2) +
  scale_color_manual(values = c("Weekday" = "#08519c", "Weekend" = "#6baed6")) +
  labs(
    title = "Average Hourly Ridership Patterns Q1",
    subtitle = "Clear commute patterns on weekdays",
    x = "Hour of Day",
    y = "Average Trips per Hour",
    color = "Day Type"
  ) +
  plotTheme

```

```{r top stations}
# Most popular origin stations
top_stations_q3 <- indego_q3 %>%
  count(start_station, start_lat, start_lon, name = "trips") %>%
  arrange(desc(trips)) %>%
  head(20)

kable(top_stations_q3, 
      caption = "Top 20 Indego Stations by Trip Origins Q3",
      format.args = list(big.mark = ",")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

top_stations_q1 <- indego_q1 %>%
  count(start_station, start_lat, start_lon, name = "trips") %>%
  arrange(desc(trips)) %>%
  head(20)

kable(top_stations_q1, 
      caption = "Top 20 Indego Stations by Trip Origins Q1",
      format.args = list(big.mark = ",")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))


```

```{r census data}
#| message: false
#| warning: false


# Get Philadelphia census tracts
philly_census <- get_acs(
  geography = "tract",
  variables = c(
    "B01003_001",  # Total population
    "B19013_001",  # Median household income
    "B08301_001",  # Total commuters
    "B08301_010",  # Commute by transit
    "B02001_002",  # White alone
    "B25077_001"   # Median home value
  ),
  state = "PA",
  county = "Philadelphia",
  year = 2022,
  geometry = TRUE,
  output = "wide"
) %>%
  rename(
    Total_Pop = B01003_001E,
    Med_Inc = B19013_001E,
    Total_Commuters = B08301_001E,
    Transit_Commuters = B08301_010E,
    White_Pop = B02001_002E,
    Med_Home_Value = B25077_001E
  ) %>%
  mutate(
    Percent_Taking_Transit = (Transit_Commuters / Total_Commuters) * 100,
    Percent_White = (White_Pop / Total_Pop) * 100
  ) %>%
  st_transform(crs = 4326)  #transforming to WGS84 for lat/lon matching, currently in NAD83

# Check the data
glimpse(philly_census)
```
```{r map philly}
# Map median income
ggplot() +
  geom_sf(data = philly_census, aes(fill = Med_Inc), color = NA) +
  scale_fill_viridis(
    option = "viridis",
    name = "Median\nIncome",
    labels = scales::dollar
  ) +
  labs(
    title = "Philadelphia Median Household Income by Census Tract",
    subtitle = "Context for understanding bike share demand patterns"
  ) +
  # Stations 
  geom_point(
    data = indego_q3,
    aes(x = start_lon, y = start_lat),
    color = "red", size = 0.25, alpha = 0.6
  ) +
  mapTheme
```

```{r join to station data}
# Create sf object for stations
stations_sf <- indego_q3 %>%
  distinct(start_station, start_lat, start_lon) %>%
  filter(!is.na(start_lat), !is.na(start_lon)) %>%
  st_as_sf(coords = c("start_lon", "start_lat"), crs = 4326)

# Spatial join to get census tract for each station
stations_census <- st_join(stations_sf, philly_census, left = TRUE) %>%
  st_drop_geometry()

# Look at the result - investigate whether all of the stations joined to census data -- according to the map above there are stations in non-residential tracts.

stations_for_map <- indego_q3 %>%
  distinct(start_station, start_lat, start_lon) %>%
  filter(!is.na(start_lat), !is.na(start_lon)) %>%
  left_join(
    stations_census %>% select(start_station, Med_Inc),
    by = "start_station"
  ) %>%
  mutate(has_census = !is.na(Med_Inc))

# Add back to trip data
indego_census_q3 <- indego_q3 %>%
  left_join(
    stations_census %>% 
      select(start_station, Med_Inc, Percent_Taking_Transit, 
             Percent_White, Total_Pop),
    by = "start_station"
  )

indego_census_q1 <- indego_q1 %>%
  left_join(
    stations_census %>% 
      select(start_station, Med_Inc, Percent_Taking_Transit, 
             Percent_White, Total_Pop),
    by = "start_station"
  )


# Prepare data for visualization
stations_for_map <- indego_q3 %>%
  distinct(start_station, start_lat, start_lon) %>%
  filter(!is.na(start_lat), !is.na(start_lon)) %>%
  left_join(
    stations_census %>% select(start_station, Med_Inc),
    by = "start_station"
  ) %>%
  mutate(has_census = !is.na(Med_Inc))

# Create the map showing problem stations
ggplot() +
  geom_sf(data = philly_census, aes(fill = Med_Inc), color = "white", size = 0.05) +
  scale_fill_viridis(
    option = "viridis",
    name = "Median\nIncome",
    labels = scales::dollar,
    na.value = "grey90"
  ) +
  # Stations with census data (small grey dots)
  geom_point(
    data = stations_for_map %>% filter(has_census),
    aes(x = start_lon, y = start_lat),
    color = "grey30", size = 1, alpha = 0.6
  ) +
  # Stations WITHOUT census data (red X marks the spot)
  geom_point(
    data = stations_for_map %>% filter(!has_census),
    aes(x = start_lon, y = start_lat),
    color = "red", size = 1, shape = 4, stroke = 1.5
  ) +
  labs(
    title = "Philadelphia Median Household Income by Census Tract",
    subtitle = "Indego stations shown (RED = no census data match)",
    caption = "Red X marks indicate stations that didn't join to census tracts"
  ) +
  mapTheme
```

```{r valid stations}
# Identify which stations to keep
valid_stations <- stations_census %>%
  filter(!is.na(Med_Inc)) %>%
  pull(start_station)

# Filter trip data to valid stations only
indego_census <- indego_q3 %>%
  filter(start_station %in% valid_stations) %>%
  left_join(
    stations_census %>% 
      select(start_station, Med_Inc, Percent_Taking_Transit, 
             Percent_White, Total_Pop),
    by = "start_station"
  )
```

```{r weather}
# Get weather from Philadelphia International Airport (KPHL)
# This covers Q3 2025: July 1 - September 3o
weather_data_q3 <- riem_measures(
  station = "PHL",  # Philadelphia International Airport
  date_start = "2025-07-01",
  date_end = "2025-09-30"
)

# Process weather data
weather_processed_q3 <- weather_data_q3 %>%
  mutate(
    interval60 = floor_date(valid, unit = "hour"),
    Temperature = tmpf,  # Temperature in Fahrenheit
    Precipitation = ifelse(is.na(p01i), 0, p01i),  # Hourly precip in inches
    Wind_Speed = sknt  # Wind speed in knots
  ) %>%
  select(interval60, Temperature, Precipitation, Wind_Speed) %>%
  distinct(interval60, .keep_all = TRUE)

# Check for missing hours and interpolate if needed
weather_complete_q3 <- weather_processed_q3 %>%
  complete(interval60 = seq(min(interval60), max(interval60), by = "hour")) %>%
  fill(Temperature, Precipitation, Wind_Speed, .direction = "down")

# Look at the weather
summary(weather_complete_q3 %>% select(Temperature, Precipitation, Wind_Speed))

# This covers Q1 2025: January 1 - March 31
weather_data_q1 <- riem_measures(
  station = "PHL",  # Philadelphia International Airport
  date_start = "2025-01-01",
  date_end = "2025-03-31"
)

# Process weather data
weather_processed_q1 <- weather_data_q1 %>%
  mutate(
    interval60 = floor_date(valid, unit = "hour"),
    Temperature = tmpf,  # Temperature in Fahrenheit
    Precipitation = ifelse(is.na(p01i), 0, p01i),  # Hourly precip in inches
    Wind_Speed = sknt  # Wind speed in knots
  ) %>%
  select(interval60, Temperature, Precipitation, Wind_Speed) %>%
  distinct(interval60, .keep_all = TRUE)

# Check for missing hours and interpolate if needed
weather_complete_q1 <- weather_processed_q1 %>%
  complete(interval60 = seq(min(interval60), max(interval60), by = "hour")) %>%
  fill(Temperature, Precipitation, Wind_Speed, .direction = "down")

# Look at the weather
summary(weather_complete_q1 %>% select(Temperature, Precipitation, Wind_Speed))
```

```{r visualize weather}
#| message: false
#| warning: false
#| 
ggplot(weather_complete_q3, aes(x = interval60, y = Temperature)) +
  geom_line(color = "#3182bd", alpha = 0.7) +
  geom_smooth(se = FALSE, color = "red") +
  labs(
    title = "Philadelphia Temperature - Q3 2025",
    subtitle = "Summer to early fall transition",
    x = "Date",
    y = "Temperature (°F)"
  ) +
  plotTheme

ggplot(weather_complete_q1, aes(x = interval60, y = Temperature)) +
  geom_line(color = "#3182bd", alpha = 0.7) +
  geom_smooth(se = FALSE, color = "red") +
  labs(
    title = "Philadelphia Temperature - Q1 2025",
    subtitle = "Winter to early spring transition",
    x = "Date",
    y = "Temperature (°F)"
  ) +
  plotTheme

```

```{r aggregate data}
#| message: false
#| warning: false

# Count trips by station-hour
trips_panel_q3 <- indego_census_q3 %>%
  group_by(interval60, start_station, start_lat, start_lon,
           Med_Inc, Percent_Taking_Transit, Percent_White, Total_Pop) %>%
  summarize(Trip_Count = n()) %>%
  ungroup()

trips_panel_q1 <- indego_census_q1 %>%
  group_by(interval60, start_station, start_lat, start_lon,
           Med_Inc, Percent_Taking_Transit, Percent_White, Total_Pop) %>%
  summarize(Trip_Count = n()) %>%
  ungroup()

# How many station-hour observations?
nrow(trips_panel_q3)

# How many unique stations?
length(unique(trips_panel_q3$start_station))

# How many unique hours?
length(unique(trips_panel_q3$interval60))

hist(trips_panel_q3$Trip_Count)
```

```{r complete panel}
# Calculate expected panel size
n_stations_q1 <- length(unique(trips_panel_q1$start_station))
n_hours_q1 <- length(unique(trips_panel_q1$interval60))
expected_rows_q1 <- n_stations_q1 * n_hours_q1

cat("Expected panel rows:", format(expected_rows_q1, big.mark = ","), "\n")
cat("Current rows:", format(nrow(trips_panel_q1), big.mark = ","), "\n")
cat("Missing rows:", format(expected_rows_q1 - nrow(trips_panel_q1), big.mark = ","), "\n")

n_stations_q3 <- length(unique(trips_panel_q3$start_station))
n_hours_q3 <- length(unique(trips_panel_q3$interval60))
expected_rows_q3 <- n_stations_q3 * n_hours_q3

cat("Expected panel rows:", format(expected_rows_q3, big.mark = ","), "\n")
cat("Current rows:", format(nrow(trips_panel_q3), big.mark = ","), "\n")
cat("Missing rows:", format(expected_rows_q3 - nrow(trips_panel_q3), big.mark = ","), "\n")

# Create complete panel Q3
study_panel_q3 <- expand_grid(
  interval60 = unique(trips_panel_q3$interval60),
  start_station = unique(trips_panel_q3$start_station)
) %>%
  # Join trip counts
  left_join(trips_panel_q3 %>% 
              select(interval60, start_station, Trip_Count),
            by = c("interval60", "start_station")) %>%
  # Replace NA trip counts with 0
  mutate(Trip_Count = replace_na(Trip_Count, 0))

glimpse(study_panel_q3)
hist(study_panel_q3$Trip_Count)

# Create complete panel Q1
study_panel_q1 <- expand_grid(
  interval60 = unique(trips_panel_q1$interval60),
  start_station = unique(trips_panel_q1$start_station)
) %>%
  left_join(trips_panel_q1 %>% 
              select(interval60, start_station, Trip_Count),
            by = c("interval60", "start_station")) %>%
  mutate(Trip_Count = replace_na(Trip_Count, 0))

glimpse(study_panel_q1)
hist(study_panel_q1$Trip_Count)

# Fill in station attributes (they're the same for all hours)
station_attributes_q3 <- trips_panel_q3 %>%
  group_by(start_station) %>%
  summarize(
    start_lat = first(start_lat),
    start_lon = first(start_lon),
    Med_Inc = first(Med_Inc),
    Percent_Taking_Transit = first(Percent_Taking_Transit),
    Percent_White = first(Percent_White),
    Total_Pop = first(Total_Pop)
  )
station_attributes_q1 <- trips_panel_q1 %>%
  group_by(start_station) %>%
  summarize(
    start_lat = first(start_lat),
    start_lon = first(start_lon),
    Med_Inc = first(Med_Inc),
    Percent_Taking_Transit = first(Percent_Taking_Transit),
    Percent_White = first(Percent_White),
    Total_Pop = first(Total_Pop)
  )

glimpse(station_attributes_q3)

study_panel_q1 <- study_panel_q1 %>%
  left_join(station_attributes_q1, by = "start_station")
study_panel_q3 <- study_panel_q3 %>%
  left_join(station_attributes_q3, by = "start_station")

glimpse(study_panel_q1)
glimpse(study_panel_q3)

# Verify we have complete panel
cat("Complete panel rows:", format(nrow(study_panel_q3), big.mark = ","), "\n")
cat("Complete panel rows:", format(nrow(study_panel_q1), big.mark = ","), "\n")

```

```{r add time features}
study_panel_q3 <- study_panel_q3 %>%
  mutate(
    week = week(interval60),
    month = month(interval60, label = TRUE),
    dotw = wday(interval60, label = TRUE),
    hour = hour(interval60),
    date = as.Date(interval60),
    weekend = ifelse(dotw %in% c("Sat", "Sun"), 1, 0),
    rush_hour = ifelse(hour %in% c(7, 8, 9, 16, 17, 18), 1, 0)
  )

glimpse(study_panel_q3)

study_panel_q1 <- study_panel_q1 %>%
  mutate(
    week = week(interval60),
    month = month(interval60, label = TRUE),
    dotw = wday(interval60, label = TRUE),
    hour = hour(interval60),
    date = as.Date(interval60),
    weekend = ifelse(dotw %in% c("Sat", "Sun"), 1, 0),
    rush_hour = ifelse(hour %in% c(7, 8, 9, 16, 17, 18), 1, 0)
  )

glimpse(study_panel_q1)
```

```{r join weather}
study_panel_q3 <- study_panel_q3 %>%
  left_join(weather_complete_q3, by = "interval60")

glimpse(study_panel_q3)

study_panel_q1 <- study_panel_q1 %>%
  left_join(weather_complete_q1, by = "interval60")

glimpse(study_panel_q1)

# Check for missing values
summary(study_panel_q3 %>% select(Trip_Count, Temperature, Precipitation))
summary(study_panel_q1 %>% select(Trip_Count, Temperature, Precipitation))

```

```{r create lags}
# Sort by station and time
study_panel_q3 <- study_panel_q3 %>%
  arrange(start_station, interval60)

# Create lag variables WITHIN each station
study_panel_q3 <- study_panel_q3 %>%
  group_by(start_station) %>%
  mutate(
    lag1Hour = lag(Trip_Count, 1),
    lag2Hours = lag(Trip_Count, 2),
    lag3Hours = lag(Trip_Count, 3),
    lag12Hours = lag(Trip_Count, 12),
    lag1day = lag(Trip_Count, 24)
  ) %>%
  ungroup()

# Sort by station and time
study_panel_q1 <- study_panel_q1 %>%
  arrange(start_station, interval60)

# Create lag variables WITHIN each station
study_panel_q1 <- study_panel_q1 %>%
  group_by(start_station) %>%
  mutate(
    lag1Hour = lag(Trip_Count, 1),
    lag2Hours = lag(Trip_Count, 2),
    lag3Hours = lag(Trip_Count, 3),
    lag12Hours = lag(Trip_Count, 12),
    lag1day = lag(Trip_Count, 24)
  ) %>%
  ungroup()

# Remove rows with NA lags (first 24 hours for each station)
study_panel_complete_q3 <- study_panel_q3 %>%
  filter(!is.na(lag1day))
study_panel_complete_q1 <- study_panel_q1 %>%
  filter(!is.na(lag1day))

cat("Rows after removing NA lags:", format(nrow(study_panel_complete_q1), big.mark = ","), "\n")
cat("Rows after removing NA lags:", format(nrow(study_panel_complete_q3), big.mark = ","), "\n")

```

```{r lag visualizations}
# Sample one station to visualize
example_station <- study_panel_complete_q3 %>%
  filter(start_station == 3188) %>%
  head(168)  # One week

# Plot actual vs lagged demand
ggplot(example_station, aes(x = interval60)) +
  geom_line(aes(y = Trip_Count, color = "Current"), linewidth = 1) +
  geom_line(aes(y = lag1Hour, color = "1 Hour Ago"), linewidth = 1, alpha = 0.7) +
  geom_line(aes(y = lag1day, color = "24 Hours Ago"), linewidth = 1, alpha = 0.7) +
  scale_color_manual(values = c(
    "Current" = "#08519c",
    "1 Hour Ago" = "#3182bd",
    "24 Hours Ago" = "#6baed6"
  )) +
  labs(
    title = "Temporal Lag Patterns at One Station",
    subtitle = "Past demand predicts future demand",
    x = "Date-Time",
    y = "Trip Count",
    color = "Time Period"
  ) +
  plotTheme
```

```{r temporal test train}
# Split by week
# Q3 has weeks  26-39 (Jul-Sep)
# Train on weeks 26-35 (Jul 1- early September)
# Test on weeks 36-39 (rest of September)

# Split by week
# Q1 has weeks 1-13 (Jan-Mar)
# Train on weeks 1-9 (Jan 1 - early March)
# Test on weeks 10-13 (rest of March)

# Which stations have trips in BOTH early and late periods?
early_stations_q3 <- study_panel_complete_q3 %>%
  filter(week < 36) %>%
  filter(Trip_Count > 0) %>%
  distinct(start_station) %>%
  pull(start_station)

late_stations_q3 <- study_panel_complete_q3 %>%
  filter(week >= 36) %>%
  filter(Trip_Count > 0) %>%
  distinct(start_station) %>%
  pull(start_station)

early_stations_q1 <- study_panel_complete_q1 %>%
  filter(week < 10) %>%
  filter(Trip_Count > 0) %>%
  distinct(start_station) %>%
  pull(start_station)

late_stations_q1 <- study_panel_complete_q1 %>%
  filter(week >= 10) %>%
  filter(Trip_Count > 0) %>%
  distinct(start_station) %>%
  pull(start_station)

# Keep only stations that appear in BOTH periods
common_stations_q3 <- intersect(early_stations_q3, late_stations_q3)
common_stations_q1 <- intersect(early_stations_q1, late_stations_q1)

# Filter panel to only common stations
study_panel_complete_q3 <- study_panel_complete_q3 %>%
  filter(start_station %in% common_stations_q3)

study_panel_complete_q1 <- study_panel_complete_q1 %>%
  filter(start_station %in% common_stations_q1)

# Create train/test split
train_q3 <- study_panel_complete_q3 %>%
  filter(week < 36)

test_q3 <- study_panel_complete_q3 %>%
  filter(week >= 36)

train_q1 <- study_panel_complete_q1 %>%
  filter(week < 10)

test_q1 <- study_panel_complete_q1 %>%
  filter(week >= 10)

# Convert start_station to factor in training data first,
# then align test levels to training levels so predict() doesn't error
# Convert to factor FIRST
train_q3 <- train_q3 %>%
  mutate(start_station = factor(start_station))

train_q1 <- train_q1 %>%
  mutate(start_station = factor(start_station))

train_q3 <- train_q3 %>% droplevels()
train_q1 <- train_q1 %>% droplevels()

cat("Training observations:", format(nrow(train_q3), big.mark = ","), "\n")
cat("Testing observations:", format(nrow(test_q3), big.mark = ","), "\n")
cat("Training date range:", min(train_q3$date), "to", max(train_q3$date), "\n")
cat("Testing date range:", min(test_q3$date), "to", max(test_q3$date), "\n")

cat("Training observations:", format(nrow(train_q1), big.mark = ","), "\n")
cat("Testing observations:", format(nrow(test_q1), big.mark = ","), "\n")
cat("Training date range:", min(train_q1$date), "to", max(train_q1$date), "\n")
cat("Testing date range:", min(test_q1$date), "to", max(test_q1$date), "\n")


```

```{r model 1}

# Create day of week factor with treatment (dummy) coding
train_q3 <- train_q3 %>%
  mutate(dotw_simple = factor(dotw, levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")))

# Set contrasts to treatment coding (dummy variables)
contrasts(train_q3$dotw_simple) <- contr.treatment(7)

# Now run the model
model1_q3 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation,
  data = train_q3
)

summary(model1_q3)

# Create day of week factor with treatment (dummy) coding
train_q1 <- train_q1 %>%
  mutate(dotw_simple = factor(dotw, levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")))

# Set contrasts to treatment coding (dummy variables)
contrasts(train_q1$dotw_simple) <- contr.treatment(7)

# Now run the model
model1_q1 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation,
  data = train_q1
)

summary(model1_q1)
```

```{r model 2}
model2_q3 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day,
  data = train_q3
)
summary(model2_q3)

model2_q1 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day,
  data = train_q1
)
summary(model2_q1)


```

```{r model 3}
model3_q3 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day +
    Med_Inc + Percent_Taking_Transit + Percent_White,
  data = train_q3
)

summary(model3_q3)

model3_q1 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day +
    Med_Inc + Percent_Taking_Transit + Percent_White,
  data = train_q1
)

summary(model3_q1)


```

```{r model 4}
model4_q3 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day +
    Med_Inc + Percent_Taking_Transit + Percent_White +
    start_station,
  data = train_q3
)
cat("Model 4 R-squared:", summary(model4_q3)$r.squared, "\n")
cat("Model 4 Adj R-squared:", summary(model4_q3)$adj.r.squared, "\n")

model4_q1 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day +
    Med_Inc + Percent_Taking_Transit + Percent_White +
    start_station,
  data = train_q1
)
cat("Model 4 R-squared:", summary(model4_q1)$r.squared, "\n")
cat("Model 4 Adj R-squared:", summary(model4_q1)$adj.r.squared, "\n")



```

```{r model 5}
model5_q3 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day + rush_hour + as.factor(month) +
    Med_Inc + Percent_Taking_Transit + Percent_White +
    start_station +
    rush_hour * weekend,  # Rush hour effects different on weekends
  data = train_q3
)

cat("Model 5 R-squared:", summary(model5_q3)$r.squared, "\n")
cat("Model 5 Adj R-squared:", summary(model5_q3)$adj.r.squared, "\n")

model5_q1 <- lm(
  Trip_Count ~ as.factor(hour) + dotw_simple + Temperature + Precipitation +
    lag1Hour + lag3Hours + lag1day + rush_hour + as.factor(month) +
    Med_Inc + Percent_Taking_Transit + Percent_White +
    start_station +
    rush_hour * weekend,  # Rush hour effects different on weekends
  data = train_q1
)

cat("Model 5 R-squared:", summary(model5_q1)$r.squared, "\n")
cat("Model 5 Adj R-squared:", summary(model5_q1)$adj.r.squared, "\n")

```

```{r mae}
# Get predictions on test set

# Create day of week factor with treatment (dummy) coding
test_q3 <- test_q3 %>%
  mutate(dotw_simple = factor(dotw, levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")))

# Set contrasts to treatment coding (dummy variables)
contrasts(test_q3$dotw_simple) <- contr.treatment(7)

test_q3$start_station <- factor(
  test_q3$start_station,
  levels = model4_q3$xlevels$start_station
)

test_q1$start_station <- factor(
  test_q1$start_station,
  levels = model4_q1$xlevels$start_station
)

test_q3 <- test_q3 %>% droplevels()
test_q1 <- test_q1 %>% droplevels()

test_q3 <- test_q3 %>%
  mutate(
    pred1 = predict(model1_q3, newdata = test_q3),
    pred2 = predict(model2_q3, newdata = test_q3),
    pred3 = predict(model3_q3, newdata = test_q3),
    pred4 = predict(model4_q3, newdata = test_q3),
    pred5 = predict(model5_q3, newdata = test_q3)
  )

# Calculate MAE for each model
mae_results_q3 <- data.frame(
  Model = c(
    "1. Time + Weather",
    "2. + Temporal Lags",
    "3. + Demographics",
    "4. + Station FE",
    "5. + Rush Hour Interaction"
  ),
  MAE = c(
    mean(abs(test_q3$Trip_Count - test_q3$pred1), na.rm = TRUE),
    mean(abs(test_q3$Trip_Count - test_q3$pred2), na.rm = TRUE),
    mean(abs(test_q3$Trip_Count - test_q3$pred3), na.rm = TRUE),
    mean(abs(test_q3$Trip_Count - test_q3$pred4), na.rm = TRUE),
    mean(abs(test_q3$Trip_Count - test_q3$pred5), na.rm = TRUE)
  )
)

kable(mae_results_q3, 
      digits = 2,
      caption = "Mean Absolute Error by Model, Q3 (Test Set)",
      col.names = c("Model", "MAE (trips)")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))


# Create day of week factor with treatment (dummy) coding
test_q1 <- test_q1 %>%
  mutate(dotw_simple = factor(dotw, levels = c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")))

# Set contrasts to treatment coding (dummy variables)
contrasts(test_q1$dotw_simple) <- contr.treatment(7)

test_q1 <- test_q1 %>%
  mutate(
    pred1 = predict(model1_q1, newdata = test_q1),
    pred2 = predict(model2_q1, newdata = test_q1),
    pred3 = predict(model3_q1, newdata = test_q1),
    pred4 = predict(model4_q1, newdata = test_q1),
    pred5 = predict(model5_q1, newdata = test_q1)
  )

# Calculate MAE for each model
mae_results_q1 <- data.frame(
  Model = c(
    "1. Time + Weather",
    "2. + Temporal Lags",
    "3. + Demographics",
    "4. + Station FE",
    "5. + Rush Hour Interaction"
  ),
  MAE = c(
    mean(abs(test_q1$Trip_Count - test_q1$pred1), na.rm = TRUE),
    mean(abs(test_q1$Trip_Count - test_q1$pred2), na.rm = TRUE),
    mean(abs(test_q1$Trip_Count - test_q1$pred3), na.rm = TRUE),
    mean(abs(test_q1$Trip_Count - test_q1$pred4), na.rm = TRUE),
    mean(abs(test_q1$Trip_Count - test_q1$pred5), na.rm = TRUE)
  )
)

kable(mae_results_q1, 
      digits = 2,
      caption = "Mean Absolute Error by Model, q1 (Test Set)",
      col.names = c("Model", "MAE (trips)")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

```

```{r visualize comparisons}
ggplot(mae_results_q3, aes(x = reorder(Model, -MAE), y = MAE)) +
  geom_col(fill = "#3182bd", alpha = 0.8) +
  geom_text(aes(label = round(MAE, 2)), vjust = -0.5) +
  labs(
    title = "Model Performance Comparison Q3",
    subtitle = "Lower MAE = Better Predictions",
    x = "Model",
    y = "Mean Absolute Error (trips)"
  ) +
  plotTheme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggplot(mae_results_q1, aes(x = reorder(Model, -MAE), y = MAE)) +
  geom_col(fill = "#3182bd", alpha = 0.8) +
  geom_text(aes(label = round(MAE, 2)), vjust = -0.5) +
  labs(
    title = "Model Performance Comparison Q1",
    subtitle = "Lower MAE = Better Predictions",
    x = "Model",
    y = "Mean Absolute Error (trips)"
  ) +
  plotTheme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r observations vs predictions}
#| message: false
#| warning: false

test_q3 <- test_q3 %>%
  mutate(
    error = Trip_Count - pred2,
    abs_error = abs(error),
    time_of_day = case_when(
      hour < 7 ~ "Overnight",
      hour >= 7 & hour < 10 ~ "AM Rush",
      hour >= 10 & hour < 15 ~ "Mid-Day",
      hour >= 15 & hour <= 18 ~ "PM Rush",
      hour > 18 ~ "Evening"
    )
  )

# Scatter plot by time and day type
ggplot(test_q3, aes(x = Trip_Count, y = pred2)) +
  geom_point(alpha = 0.2, color = "#3182bd") +
  geom_abline(slope = 1, intercept = 0, color = "red", linewidth = 1) +
  geom_smooth(method = "lm", se = FALSE, color = "darkgreen") +
  facet_grid(weekend ~ time_of_day) +
  labs(
    title = "Observed vs. Predicted Bike Trips Q3",
    subtitle = "Model 2 performance by time period Q3",
    x = "Observed Trips",
    y = "Predicted Trips",
    caption = "Red line = perfect predictions; Green line = actual model fit"
  ) +
  plotTheme

test_q1 <- test_q1 %>%
  mutate(
    error = Trip_Count - pred2,
    abs_error = abs(error),
    time_of_day = case_when(
      hour < 7 ~ "Overnight",
      hour >= 7 & hour < 10 ~ "AM Rush",
      hour >= 10 & hour < 15 ~ "Mid-Day",
      hour >= 15 & hour <= 18 ~ "PM Rush",
      hour > 18 ~ "Evening"
    )
  )

# Scatter plot by time and day type
ggplot(test_q1, aes(x = Trip_Count, y = pred2)) +
  geom_point(alpha = 0.2, color = "#3182bd") +
  geom_abline(slope = 1, intercept = 0, color = "red", linewidth = 1) +
  geom_smooth(method = "lm", se = FALSE, color = "darkgreen") +
  facet_grid(weekend ~ time_of_day) +
  labs(
    title = "Observed vs. Predicted Bike Trips q1",
    subtitle = "Model 2 performance by time period q1",
    x = "Observed Trips",
    y = "Predicted Trips",
    caption = "Red line = perfect predictions; Green line = actual model fit"
  ) +
  plotTheme

```

```{r spatial errors}
# Calculate MAE by station
station_errors_q3 <- test_q3 %>%
  group_by(start_station, start_lat, start_lon) %>%
  summarize(
    MAE = mean(abs_error, na.rm = TRUE),
    avg_demand = mean(Trip_Count, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(!is.na(start_lat), !is.na(start_lon))

station_errors_q1 <- test_q1 %>%
  group_by(start_station, start_lat, start_lon) %>%
  summarize(
    MAE = mean(abs_error, na.rm = TRUE),
    avg_demand = mean(Trip_Count, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(!is.na(start_lat), !is.na(start_lon))


## Create Two Maps Side-by-Side with Proper Legends (sorry these maps are ugly)

# Calculate station errors
station_errors_q3 <- test_q3 %>%
  filter(!is.na(pred2)) %>%
  group_by(start_station, start_lat, start_lon) %>%
  summarize(
    MAE = mean(abs(Trip_Count - pred2), na.rm = TRUE),
    avg_demand = mean(Trip_Count, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(!is.na(start_lat), !is.na(start_lon))

station_errors_q1 <- test_q1 %>%
  filter(!is.na(pred2)) %>%
  group_by(start_station, start_lat, start_lon) %>%
  summarize(
    MAE = mean(abs(Trip_Count - pred2), na.rm = TRUE),
    avg_demand = mean(Trip_Count, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(!is.na(start_lat), !is.na(start_lon))

# Map 1: Prediction Errors
p1_q3 <- ggplot() +
  geom_sf(data = philly_census, fill = "grey95", color = "white", size = 0.2) +
  geom_point(
    data = station_errors_q3,
    aes(x = start_lon, y = start_lat, color = MAE),
    size = 3.5,
    alpha = 0.7
  ) +
  scale_color_viridis(
    option = "plasma",
    name = "MAE\n(trips)",
    direction = -1,
    breaks = c(0.5, 1.0, 1.5),  # Fewer, cleaner breaks
    labels = c("0.5", "1.0", "1.5")
  ) +
  labs(title = "Prediction Errors Q3",
       subtitle = "Higher in Center City") +
  mapTheme +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 9),
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 10)
  ) +
  guides(color = guide_colorbar(
    barwidth = 1.5,
    barheight = 12,
    title.position = "top",
    title.hjust = 0.5
  ))

p1_q1 <- ggplot() +
  geom_sf(data = philly_census, fill = "grey95", color = "white", size = 0.2) +
  geom_point(
    data = station_errors_q1,
    aes(x = start_lon, y = start_lat, color = MAE),
    size = 3.5,
    alpha = 0.7
  ) +
  scale_color_viridis(
    option = "plasma",
    name = "MAE\n(trips)",
    direction = -1,
    breaks = c(0.5, 1.0, 1.5),  # Fewer, cleaner breaks
    labels = c("0.5", "1.0", "1.5")
  ) +
  labs(title = "Prediction Errors Q1",
       subtitle = "Higher in Center City") +
  mapTheme +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 9),
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 10)
  ) +
  guides(color = guide_colorbar(
    barwidth = 1.5,
    barheight = 12,
    title.position = "top",
    title.hjust = 0.5
  ))

# Map 2: Average Demand
p2_q3 <- ggplot() +
  geom_sf(data = philly_census, fill = "grey95", color = "white", size = 0.2) +
  geom_point(
    data = station_errors_q3,
    aes(x = start_lon, y = start_lat, color = avg_demand),
    size = 3.5,
    alpha = 0.7
  ) +
  scale_color_viridis(
    option = "viridis",
    name = "Avg\nDemand",
    direction = -1,
    breaks = c(0.5, 1.0, 1.5, 2.0, 2.5),  # Clear breaks
    labels = c("0.5", "1.0", "1.5", "2.0", "2.5")
  ) +
  labs(title = "Average Demand Q3",
       subtitle = "Trips per station-hour") +
  mapTheme +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 9),
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 10)
  ) +
  guides(color = guide_colorbar(
    barwidth = 1.5,
    barheight = 12,
    title.position = "top",
    title.hjust = 0.5
  ))

p2_q1 <- ggplot() +
  geom_sf(data = philly_census, fill = "grey95", color = "white", size = 0.2) +
  geom_point(
    data = station_errors_q1,
    aes(x = start_lon, y = start_lat, color = avg_demand),
    size = 3.5,
    alpha = 0.7
  ) +
  scale_color_viridis(
    option = "viridis",
    name = "Avg\nDemand",
    direction = -1,
    breaks = c(0.5, 1.0, 1.5, 2.0, 2.5),  # Clear breaks
    labels = c("0.5", "1.0", "1.5", "2.0", "2.5")
  ) +
  labs(title = "Average Demand q1",
       subtitle = "Trips per station-hour") +
  mapTheme +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 9),
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 10)
  ) +
  guides(color = guide_colorbar(
    barwidth = 1.5,
    barheight = 12,
    title.position = "top",
    title.hjust = 0.5
  ))



# Combine
(p1_q3 | p2_q3)
(p1_q1 | p2_q1)
```

```{r temporal errors}
# MAE by time of day and day type
temporal_errors_q3 <- test_q3 %>%
  group_by(time_of_day, weekend) %>%
  summarize(
    MAE = mean(abs_error, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(day_type = ifelse(weekend == 1, "Weekend", "Weekday"))

ggplot(temporal_errors_q3, aes(x = time_of_day, y = MAE, fill = day_type)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Weekday" = "#08519c", "Weekend" = "#6baed6")) +
  labs(
    title = "Prediction Errors by Time Period Q3",
    subtitle = "When is the model struggling most?",
    x = "Time of Day",
    y = "Mean Absolute Error (trips)",
    fill = "Day Type"
  ) +
  plotTheme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

temporal_errors_q1 <- test_q1 %>%
  group_by(time_of_day, weekend) %>%
  summarize(
    MAE = mean(abs_error, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(day_type = ifelse(weekend == 1, "Weekend", "Weekday"))

ggplot(temporal_errors_q1, aes(x = time_of_day, y = MAE, fill = day_type)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Weekday" = "#08519c", "Weekend" = "#6baed6")) +
  labs(
    title = "Prediction Errors by Time Period Q1",
    subtitle = "When is the model struggling most?",
    x = "Time of Day",
    y = "Mean Absolute Error (trips)",
    fill = "Day Type"
  ) +
  plotTheme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r demographic errors}
#| message: false
#| warning: false

# Join demographic data to station errors

station_errors_demo_q3 <- station_errors_q3 %>%
  mutate(start_station = as.character(start_station)) %>%
  left_join(
    station_attributes_q3 %>%
      mutate(start_station = as.character(start_station)) %>%
      select(start_station, Med_Inc, Percent_Taking_Transit, Percent_White),
    by = "start_station"
  ) %>%
  filter(!is.na(Med_Inc))

station_errors_demo_q1 <- station_errors_q1 %>%
  mutate(start_station = as.character(start_station)) %>%
  left_join(
    station_attributes_q1 %>%
      mutate(start_station = as.character(start_station)) %>%
      select(start_station, Med_Inc, Percent_Taking_Transit, Percent_White),
    by = "start_station"
  ) %>%
  filter(!is.na(Med_Inc))


# Create plots
p1_q3 <- ggplot(station_errors_demo_q3, aes(x = Med_Inc, y = MAE)) +
  geom_point(alpha = 0.5, color = "#3182bd") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  scale_x_continuous(labels = scales::dollar) +
  labs(title = "Errors vs. Median Income Q3", x = "Median Income", y = "MAE") +
  plotTheme

p2_q3 <- ggplot(station_errors_demo_q3, aes(x = Percent_Taking_Transit, y = MAE)) +
  geom_point(alpha = 0.5, color = "#3182bd") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "Errors vs. Transit Usage Q3", x = "% Taking Transit", y = "MAE") +
  plotTheme

p3_q3 <- ggplot(station_errors_demo_q3, aes(x = Percent_White, y = MAE)) +
  geom_point(alpha = 0.5, color = "#3182bd") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "Errors vs. Race Q3", x = "% White", y = "MAE") +
  plotTheme

(p1_q3 | p2_q3) / (p3_q3 | plot_spacer())

p1_q1 <- ggplot(station_errors_demo_q1, aes(x = Med_Inc, y = MAE)) +
  geom_point(alpha = 0.5, color = "#3182bd") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  scale_x_continuous(labels = scales::dollar) +
  labs(title = "Errors vs. Median Income q1", x = "Median Income", y = "MAE") +
  plotTheme

p2_q1 <- ggplot(station_errors_demo_q1, aes(x = Percent_Taking_Transit, y = MAE)) +
  geom_point(alpha = 0.5, color = "#3182bd") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "Errors vs. Transit Usage q1", x = "% Taking Transit", y = "MAE") +
  plotTheme

p3_q1 <- ggplot(station_errors_demo_q1, aes(x = Percent_White, y = MAE)) +
  geom_point(alpha = 0.5, color = "#3182bd") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "Errors vs. Race q1", x = "% White", y = "MAE") +
  plotTheme

(p1_q1 | p2_q1) / (p3_q1 | plot_spacer())

```

Potential features to consider:

Temporal features:

Holiday indicators (Memorial Day, July 4th, Labor Day)

School calendar (Penn, Drexel, Temple in session?)

Special events (concerts, sports games, conventions)

Day of month (payday effects?)

Weather features:

Feels-like temperature (wind chill/heat index)

"Perfect biking weather" indicator (60-75°F, no rain)

Precipitation forecast (not just current)

Weekend + nice weather interaction

Spatial features:

Distance to Center City

Distance to nearest university

Distance to nearest park

Points of interest nearby (restaurants, offices, bars)

Station capacity

Bike lane connectivity

Trip history features:

Rolling 7-day average demand

Same hour last week

Station "type" clustering (residential, commercial, tourist)

Implementation:

Add your features to the best model

Compare MAE before and after

Explain why you chose these features

Did they improve predictions? Where?

Try a poisson model for count data

Does this improve model fit?
